---
title: "Enhancer screen chromatin analyses"
author: "Andreas R. Gschwind"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

```{r setupDocument, include=FALSE}
# save.image("RDA/chromatin_plots.rda")
# stop()

# set output html chunk options
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

## Goal
Make main figure chromatin analyses plots.
```{r attachPackages}
library(tidyverse)
library(data.table)
library(cowplot)
library(ggpubr)
```

***

## Input data
Load all required input data
```{r loadData}
# load table with chromatin annotations for all ETPs
etp <- fread(snakemake@input[[1]], sep = ",")

# for control perturbations which were in multiple panels, select the most significant one
etp_ctrl <- etp %>% 
  filter(ctrl_perturbation == TRUE) %>% 
  group_by(response_id, grna_target) %>% 
  slice_min(p_value, n = 1, with_ties = FALSE)

# add back to results table with discovery pairs
etp <- etp %>% 
  filter(ctrl_perturbation == FALSE) %>% 
  bind_rows(etp_ctrl) %>% 
  arrange(desc(p_value))

# add pseudocount to Hi-C interaction frequencies
etp <- mutate(etp, hic_int_freq = hic_int_freq + 1)

# filter out any trans-chromosomal interactions
etp <- filter(etp, pert_chr == gene_chr)
```

***

## Chromatin features across ETP classes
Plot chromatin features of all `r sum(etp$significant, na.rm = TRUE)` hits across all classes of
perturbed target-gene pairs, including promoter and distal enhancer targeting perturbations.
```{r prepareData1}
# features to include in plot
features <- c("log_2_fold_change", "enh_signal_count_DNase-seq", "enh_signal_avg_H3K27ac",
              "enh_signal_avg_H3K4me1", "enh_signal_avg_H3K4me3", "hic_int_freq")

# extract features for significant ETPs
etp_feat <- etp %>% 
  select(response_id, grna_target, interaction_type, significant, dist_to_tss,
         enhancer_like_interaction, all_of(features)) %>% 
  pivot_longer(cols = all_of(features), names_to = "feature", values_to = "value") %>% 
  mutate(dist_to_tss = abs(dist_to_tss))

# reformat feature names for plot
etp_feat <- etp_feat %>% 
  mutate(feature = case_when(
    feature == "log_2_fold_change" ~ "Effect size",
    feature == "hic_int_freq" ~ "Hi-C 3D contact",
    TRUE ~ sub(".+_(.+)", "\\1", feature)
  ))

# convert classes to factors for correct ordering
class_levels <- c("promoter-proximal", "promoter-distal", "promoter-trans", "intragenic-proximal",
                  "intragenic-distal", "intragenic-trans", "intergenic-distal", "intergenic-trans")
etp_feat <- etp_feat %>% 
  mutate(interaction_type = factor(interaction_type, levels = rev(class_levels)))

# extract positive hits only for plots
etp_feat_sig <- filter(etp_feat, significant == TRUE)

# add effect size as separate column
etp_feat_sig <- etp_feat_sig %>% 
  filter(feature == "Effect size") %>% 
  select(response_id, grna_target, effect_size = value) %>%
  left_join(etp_feat_sig, ., by = c("response_id", "grna_target"))
```

The median feature signals are computed to use as baseline reference in subsequent plots.
```{r calcAverageFeatureValues}
# calculate average feature values across all tested ETP
avg_feature_values <- etp_feat %>% 
  group_by(feature) %>% 
  summarize(mean = mean(value, na.rm = TRUE),
            median = median(value, na.rm = TRUE))

# calculate average feature values across all negative tested elements
positive_elements <- unique(pull(filter(etp_feat, significant == TRUE), grna_target))
avg_feature_values_negatives <- etp_feat %>% 
  filter(!grna_target %in% positive_elements) %>% 
  group_by(feature) %>% 
  summarize(mean = mean(value, na.rm = TRUE),
            median = median(value, na.rm = TRUE))
```

Plot chromatin feature value distributions of positive hits across the different ETP classes for
each assay.
```{r customPlotTheme}
custom_theme <- function(...) {
  theme_bw(...) %+replace% {
    theme(
      axis.title.y = element_blank(), panel.grid = element_blank(),
      strip.background = element_blank(),
      axis.title.x = element_text(size = 9, color = "black"),
      axis.text.x = element_text(size = 9, color = "black"),
      axis.text.y = element_text(size = 9, color = "black"),
      strip.text = element_text(size = 11, color = "black")
    )
  }
}
```

Plot E-G features per class and color based in whether a given interaction is considered a real
cre-gene interaction.
```{r plotChromClassColor}
# sort table for correct plotting order for individual points
etp_feat_sig <- arrange(etp_feat_sig, enhancer_like_interaction)

# colors for true enhancers
true_enh_colors <- c("TRUE" = "#0263A3", "FALSE" = "#a9a9a9")

# make effect size plot
es_dat <- filter(etp_feat_sig, feature == "Effect size")
es_plot <- ggplot(es_dat, aes(y = interaction_type, x = value, color = enhancer_like_interaction)) +
  facet_wrap(~feature, nrow = 1, scales = "free_x") +
  geom_jitter(height = 0.2, pch = 16) +
  geom_boxplot(outlier.shape = NA, fill = NA, width = 0.5, color = "black") +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs(x = expression("log"[2] ~ "fold change"), color = "Distal CRE-gene\ninteraction") +
  scale_color_manual(values = true_enh_colors) +
  custom_theme()

# make DNase-seq plot
dnase_dat <- filter(etp_feat_sig, feature == "DNase-seq")
dnase_plot <- ggplot(dnase_dat,
                     aes(y = interaction_type, x = value, color = enhancer_like_interaction)) +
  facet_wrap(~feature, nrow = 1, scales = "free_x") +
  geom_jitter(height = 0.2, pch = 16) +
  geom_boxplot(outlier.shape = NA, fill = NA, width = 0.5, color = "black") +
  geom_vline(data = filter(avg_feature_values, feature == "DNase-seq"),
             aes(xintercept = median), linetype = "dashed") +
  labs(x = "Read depth normalized\nsignal", color = "Distal CRE-gene\ninteraction") +
  scale_color_manual(values = true_enh_colors) +
  scale_x_log10() +
  custom_theme()

# make ChIP-seq plot
chip_assays <- c("H3K27ac", "H3K4me1", "H3K4me3")
chip_data <- filter(etp_feat_sig, feature %in% chip_assays)
chip_plot <- ggplot(chip_data,
                    aes(y = interaction_type, x = value, color = enhancer_like_interaction)) +
  facet_wrap(~feature, nrow = 1, scales = "free_x") +
  geom_jitter(height = 0.2, pch = 16) +
  geom_boxplot(outlier.shape = NA, fill = NA, width = 0.5, color = "black") +
  geom_vline(data = filter(avg_feature_values, feature %in% chip_assays),
             aes(xintercept = median), linetype = "dashed") +  
  labs(x = "Fold change over control", color = "Distal CRE-gene\ninteraction") +
  scale_x_log10() +
  scale_color_manual(values = true_enh_colors) +
  custom_theme()

# make Hi-C plot
hic_data <- filter(etp_feat_sig, feature == "Hi-C 3D contact")
hic_plot <- ggplot(hic_data,
                   aes(y = interaction_type, x = value, color = enhancer_like_interaction)) +
  facet_wrap(~feature, nrow = 1, scales = "free_x") +
  geom_jitter(height = 0.2, pch = 16) +
  geom_boxplot(outlier.shape = NA, fill = NA, width = 0.5, color = "black") +
  labs(x = "Hi-C interaction\nfrequency + 1", color = "Distal CRE-gene\ninteraction") +
  scale_x_log10() +
  scale_color_manual(values = true_enh_colors) +
  custom_theme()
```

```{r arrangeChromClassColorPlot1, fig.height=4, fig.width=12.2}
# extract legend from one plot
legend <- get_legend(es_plot)

# arrange all plots into one set
panels <- plot_grid(
  es_plot + theme(legend.position = "none"),
  dnase_plot + theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(),
                     legend.position = "none"),
  chip_plot + theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(),
                    legend.position = "none"),
  hic_plot + theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(),
                   legend.position = "none"),
  nrow = 1, align = "h", rel_widths = c(0.85, 0.5, 1.4, 0.5, 0.5), scale = 0.95
)

# combine with legend to assemble entire panel
plot_grid(panels, legend, rel_widths = c(1, 0.12))
```

```{r savePlot1}
ggsave(filename = "results/plots/chromatin_assays_etp_classes.pdf", height = 4, width = 12.2,
      create.dir = TRUE)
```

***

## Distal enhancers
Plot chromatin features of distal enhancers as function of distance to TSS.
```{r prepareData2}
# make distance to TSS bin an ordered factor
distToTSS_bin_levels <- c("TSS", "1500-10k", "10k-100k", "100k-1Mb", ">1Mb")
etp <- mutate(etp, distToTSS_bin = fct_relevel(distToTSS_bin, distToTSS_bin_levels))

# filter for specified E-G pairs to include in plots
enh_etp <- etp %>%
  filter(!is.na(significant)) %>% 
  filter(interaction_type %in% c("intergenic-distal", "promoter-proximal") | enhancer_like_interaction == TRUE)

# chromatin assays to include in plot
plot_features <- c("log_2_fold_change", "enh_signal_count_DNase-seq", "enh_signal_avg_H3K27ac",
                   "enh_signal_avg_H3K4me1", "enh_signal_avg_H3K4me3")

# select any elements that were detected as positives and categorize by distance
all_bins <- c("hit TSS", "hit 1500-10k", "hit 10k-100k", "hit 100k-1Mb", "hit >1Mb", "NS")
positives <- enh_etp %>% 
  filter(significant == TRUE) %>% 
  mutate(category = paste("hit", distToTSS_bin)) %>%
  mutate(category = factor(category, levels = all_bins, ordered = TRUE) )

# select unique enhancers and features to plot
positives <- positives %>% 
  select(grna_target, significant, category, all_of(plot_features)) %>% 
  distinct()

# if an enhancer is counted in multiple categories, assign it to the closest distance to TSS bin
positives <- positives %>% 
  group_by(grna_target) %>% 
  slice_min(category, n = 1)

# filter for enhancers that are not linked to any gene and select columns for plot
negatives <- enh_etp %>% 
  filter(significant == FALSE, !grna_target %in% positives$grna_target,
         interaction_type == "intergenic-distal") %>% 
  mutate(category = factor("NS", levels = all_bins, ordered = TRUE)) %>% 
  select(grna_target, significant, category, all_of(plot_features)) %>% 
  distinct()

# combine back into one table and convert to long format for plots
plot_etp <- bind_rows(positives, negatives) %>% 
  pivot_longer(cols = all_of(plot_features),
               names_to = "column_name", values_to = "signal")

# reformat feature names for plots
plot_etp <- plot_etp %>% 
  mutate(column_name = sub("enh_signal_", "", column_name)) %>% 
  mutate(column_name = if_else(column_name == "log_2_fold_change", "LFC", column_name)) %>% 
  separate(column_name, into = c("metric", "assay"), sep = "_", fill = "right") %>% 
  mutate(assay = if_else(metric == "LFC", "Effect size", assay))
```

```{r makeChromDistPlot}
# perform statistical tests (kruskal-wallis) between categories
kw_tests <- plot_etp %>% 
  ungroup() %>% 
  compare_means(formula = signal ~ category, method = "kruskal.test", data = ., group.by = "assay") %>% 
  mutate(p.adj.plot = if_else(p.adj < 0.001, true = "< 0.001",
                              false = paste("=", round(p.adj, digits = 2))))

# get sample sizes per category group
sample_sizes <- plot_etp %>% 
  ungroup() %>% 
  select(grna_target, category) %>% 
  distinct() %>% 
  arrange(category) %>% 
  count(category, .drop = FALSE)

# set new labels (factor levels) for each category, which include sample size per category
levels <- paste0(sample_sizes$category, " (n=", sample_sizes$n, ")")
levels(plot_etp$category) <- levels

# add mock group to KW test table
kw_tests <- mutate(kw_tests, group1 = levels[[2]], group2 = levels[[3]])

# colors for enhancer categories in figures
category_colors <- structure(c("#daa520", "#0263a3", "#4299d4", "#6ebff5", "#9ad8fc", "#a9a9a9"),
                             names = levels)

# plot log2 fold change
p1 <- ggplot(filter(plot_etp, assay == "Effect size"),
             aes(x = category, y = signal, fill = category)) +
  facet_wrap(~ assay, nrow = 1) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_boxplot() +
  stat_pvalue_manual(data = filter(kw_tests, assay == "Effect size"), y.position = 2.25, size = 3,
                     label = "P adj. {p.adj.plot}", remove.bracket = TRUE, inherit.aes = FALSE) +
  #labs(y = bquote(atop("CRISPRi effect size", "(log"[2] ~ "fold change)"))) +
  labs(y = expression("log"[2] ~ "fold change")) +
  scale_fill_manual(values = category_colors) +
  scale_y_continuous(limits = c(-4, 2.5)) +
  theme_bw() +
  theme(legend.position = "none", axis.title.x = element_blank(), panel.grid = element_blank(),
        axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))

# plot DNase-seq accessibility
p2 <- ggplot(filter(plot_etp, assay == "DNase-seq"),
             aes(x = category, y = signal + 1, fill = category)) +
  facet_wrap(~ assay, nrow = 1) +
  geom_boxplot() +
  stat_pvalue_manual(data = filter(kw_tests, assay == "DNase-seq"), y.position = 15000, size = 3,
                     label = "P adj. {p.adj.plot}", remove.bracket = TRUE, inherit.aes = FALSE) +
  labs(y = paste0("Read-depth normalized", "\n", "signal")) +
  scale_fill_manual(values = category_colors) +
  scale_y_log10(limits = c(1, 20000)) +
  theme_bw() +
  theme(legend.position = "none", axis.title.x = element_blank(), panel.grid = element_blank(),
        axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))

# plot ChIP-seq signals
p3 <- ggplot(filter(plot_etp, assay %in% c("H3K27ac", "H3K4me1", "H3K4me3")),
             aes(x = category, y = signal + 0.01, fill = category)) +
  facet_wrap(~ assay, nrow = 1) +
  geom_boxplot() +
  stat_pvalue_manual(data = filter(kw_tests, assay %in% c("H3K27ac", "H3K4me1", "H3K4me3")),
                     y.position = 120, size = 3, label = "P adj. {p.adj.plot}",
                     remove.bracket = TRUE, inherit.aes = FALSE) +
  labs(y = "FC over control") +
  scale_fill_manual(values = category_colors) +
  scale_y_log10(limits = c( 0.01, 150)) +
  theme_bw() +
  theme(legend.position = "none", axis.title.x = element_blank(), panel.grid = element_blank(),
        axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
```

```{r arrangeChromDistPlot, fig.height=3.25, fig.width=10}
# assemble figure panel and save to pdf
plot_grid(p1, p2, p3, nrow = 1, rel_widths = c(0.42, 0.5, 1.1))
```

```{r savePlot2}
ggsave(filename = "results/plots/chromatin_assays_cres_vs_distance.pdf", height = 3.25, width = 10)
```

```{r ChromDistPWTests}
# perform individual pairwise tests
pw_tests <- compare_means(signal ~ category, group.by = "assay", data = ungroup(plot_etp),
                          method = "wilcox.test", p.adjust.method = "holm")

# split category labels in to labels and number of enhancers
pw_tests <- pw_tests %>% 
  separate(group1, into = c("group1", "n_group1"), sep = " \\(n=") %>% 
  separate(group2, into = c("group2", "n_group2"), sep = " \\(n=") %>% 
  mutate(n_group1 = sub("\\)", "", n_group1), n_group2 = sub("\\)", "", n_group2)) %>% 
  select(-.y.)

# save to text file
dir.create("results/tables", showWarnings = FALSE, recursive = TRUE)
write_csv(pw_tests, file = "results/tables/chromatin_features_distal_pairwise_tests.csv")
```

***

## 3D interaction frequency
Plot Hi-C 3D interaction frequencies of distal interactions as function of distance to TSS.
```{r prepareData3}
# add significance label for Hi-C plot
enh_etp_hic <- enh_etp %>% 
  filter(enhancer_like_interaction == TRUE | significant == FALSE) %>% 
  mutate(sig_label = if_else(enhancer_like_interaction, true = "Sig.", false = "NS")) %>% 
  add_count(sig_label) %>% 
  mutate(sig_label = paste0(sig_label, " (n=", n, ")")) %>% 
  mutate(sig_label = fct_reorder(sig_label, n)) %>% 
  select(-n)

# create new distance to TSS bin levels which include number of pairs
distToTSS_levels <- enh_etp_hic %>% 
  group_by(distToTSS_bin) %>% 
  summarize(sig = sum(significant == TRUE),
            nonsig = sum(significant == FALSE)) %>% 
  mutate(distToTSS_bin = as.character(distToTSS_bin)) %>% 
  mutate(distToTSS_label = paste0(distToTSS_bin, "\nSig.=", sig, "\nNS=", nonsig)) %>% 
  select(distToTSS_label, distToTSS_bin) %>% 
  deframe()

# change distance to TSS labels in ETP table
enh_etp_hic <- mutate(enh_etp_hic, distToTSS_bin = fct_recode(distToTSS_bin, !!!distToTSS_levels))

# set colors for significance labels which include number of pairs
sig_cols <- structure(c("#0263A3", "#a9a9a9"), names = levels(enh_etp_hic$sig_label))

# set NA interaction frequency to 0
enh_etp_hic <- replace_na(enh_etp_hic, replace = list(hic_int_freq = 0))

# calculate distance to TSS in bin increments
enh_etp_hic <- enh_etp_hic %>% 
  mutate(hic_bin_dist = abs(dist_to_tss) %/% snakemake@params$hic_res * snakemake@params$hic_res)
```

```{r plotHicDistBins, fig.height=3.5, fig.width=6}
# plot interaction frequencies per distance to TSS bin
ggplot(enh_etp_hic, aes(x = distToTSS_bin, y = hic_int_freq + 1, fill = sig_label)) +
  geom_boxplot() +
  labs(y = "Hi-C interaction frequency + 1", x = "Distance to TSS",
       fill = "Distal CRE-gene\ninteraction") +
  scale_fill_manual(values = sig_cols) +
  scale_y_log10(limits = c(1, 1000)) +
  theme_bw()
```

```{r savePlot3}
ggsave(filename = "results/plots/hic_etp_binned.pdf", height=3.5, width=6)
```

```{r plotHicDistBinsPWTests}
# perform significance tests between significant and non-significant pairs in each distance bin
pw_tests <- compare_means(hic_int_freq ~ significant, data = enh_etp_hic, group.by = "distToTSS_bin",
                          method = "wilcox.test", p.adjust.method = "holm")

# split category labels in to labels and number of enhancers
pw_tests <- pw_tests %>% 
  select(-c(.y., group1, group2)) %>% 
  separate(distToTSS_bin, into = c("distToTSS_bin", "group1", "n_group1", "group2", "n_group2"),
           sep = "\n|=")

# save to text file
write_csv(pw_tests, file = "results/tables/hic_interaction_frequency_distBins_pairwise_tests.csv")
```

Plot Hi-C interaction frequency as funciton of TSS for all pairs.
```{r plotHicDist, fig.width=7.25, fig.height=4}
# ass significance label
enh_etp_plot <- enh_etp_hic %>% 
  mutate(sig_label = if_else(significant, true = "Sig.", false = "NS")) %>% 
  add_count(sig_label) %>% 
  mutate(sig_label = paste0(sig_label, " (n=", n, ")")) %>% 
  mutate(sig_label = fct_reorder(sig_label, n)) %>% 
  select(-n)

# make simple 2 group boxplot
p1 <- ggplot(enh_etp_plot, aes(x = sig_label, y = hic_int_freq + 1, fill = sig_label)) +
  geom_boxplot() +
  labs(y = "Hi-C interaction frequency + 1", x = "") +
  scale_fill_manual(values = c("#0263A3", "#a9a9a9")) +
  scale_y_log10(limits = c(1, 1000)) +
  theme_bw() +
  theme(legend.position = "none", axis.text.x = element_text(angle = 45, hjust = 1))

# plot interaction frequencies as function of distance to TSS
p2 <- ggplot(enh_etp_plot, aes(x = abs(dist_to_tss) / 1000, y = hic_int_freq + 1, color = sig_label)) +
  geom_point(data = filter(enh_etp_plot, significant == FALSE)) +
  geom_point(data = filter(enh_etp_plot, significant == TRUE)) +
  geom_smooth(data = filter(enh_etp_plot, significant == FALSE), se = FALSE, color = "gray40") +
  geom_smooth(data = filter(enh_etp_plot, significant == TRUE), se = FALSE, color = "#0263A3") +
  labs(y = "Hi-C interaction frequency + 1", x = "Distance to TSS (kb)",
       color = "Distal CRE-gene\ninteraction") +
  scale_color_manual(values = c("#a9a9a9", "#0263A3")) +
  scale_x_log10(limits = c(1, NA)) +
  scale_y_log10(limits = c(1, 1000)) +
  theme_bw() +
  theme(axis.title.y = element_blank(),
        axis.text.y = element_blank())

# arrange into one figure and save to pdf file
plot_grid(p1, p2, nrow = 1, rel_widths = c(0.35, 1), align = "h")
```

```{r savePlot4}
ggsave(filename = "results/plots/hic_etp_vs_distance.pdf", width=7.25, height=4)
```

```{r plotHicDist1Mb, fig.width=7.25, fig.height=4}
# only look at results within 1Mb (typical enhancer range)
enh_etp_plot <- enh_etp_hic %>% 
  filter(abs(dist_to_tss) <= 1e6) %>% 
  mutate(sig_label = if_else(significant, true = "Sig.", false = "NS")) %>% 
  add_count(sig_label) %>% 
  mutate(sig_label = paste0(sig_label, " (n=", n, ")")) %>% 
  mutate(sig_label = fct_reorder(sig_label, n)) %>% 
  select(-n)

# make simple 2 group boxplot
p1 <- ggplot(enh_etp_plot, aes(x = sig_label, y = hic_int_freq + 1, fill = sig_label)) +
  geom_boxplot() +
  labs(y = "Hi-C interaction frequency + 1", x = "") +
  scale_fill_manual(values = c("#0263A3", "#a9a9a9")) +
  scale_y_log10(limits = c(1, 1000)) +
  theme_bw() +
  theme(legend.position = "none", axis.text.x = element_text(angle = 45, hjust = 1))

# plot interaction frequencies as function of distance to TSS
p2 <- ggplot(enh_etp_plot, aes(x = abs(dist_to_tss) / 1000, y = hic_int_freq + 1, color = sig_label)) +
  geom_point(data = filter(enh_etp_plot, significant == FALSE)) +
  geom_point(data = filter(enh_etp_plot, significant == TRUE)) +
  geom_smooth(data = filter(enh_etp_plot, significant == FALSE), se = FALSE, color = "gray40") +
  geom_smooth(data = filter(enh_etp_plot, significant == TRUE), se = FALSE, color = "#0263A3") +
  labs(y = "Hi-C interaction frequency + 1", x = "Distance to TSS (kb)",
       color = "Distal CRE-gene\ninteraction") +
  scale_color_manual(values = c("#a9a9a9", "#0263A3")) +
  scale_x_log10(limits = c(1, NA)) +
  scale_y_log10(limits = c(1, 1000)) +
  theme_bw() +
  theme(axis.title.y = element_blank(),
        axis.text.y = element_blank())

# arrange into one figure and save to pdf file
plot_grid(p1, p2, nrow = 1, rel_widths = c(0.35, 1), align = "h")
```

```{r savePlot5}
ggsave(filename = "results/plots/hic_etp_vs_distance_1mb.pdf", width=7.25, height=4)
```

***

## Session information
This document was produced using following packages:
```{r sessionInfo}
sessionInfo()
```
