---
title: "Enhancer screen chromatin analyses"
author: "Andreas R. Gschwind"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
editor_options: 
  chunk_output_type: console
---

```{r setupDocument, include=FALSE}
# save.image("RDA/chromatin_plots.rda")
# stop()

# set output html chunk options
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

## Goal
Make main figure chromatin analyses plots.
```{r attachPackages}
library(tidyverse)
library(cowplot)
library(ggpubr)
```

***

## Input data
Load all required input data
```{r loadData}
# load table with chromatin annotations for all ETPs
etp <- read_csv("/g/steinmetz/project/otar_2063/enhancer_screen/interpretation/chromatin_analyses/results/results_df_with_promoterC_annotated_allFeatures_transHits.csv")

# for control perturbations which were in multiple panels, select the most significant one
etp_ctrl <- etp %>% 
  filter(ctrl_perturbation == TRUE) %>% 
  group_by(response_id, grna_target) %>% 
  slice_min(p_value, n = 1, with_ties = FALSE)

# add back to results table with discovery pairs
etp <- etp %>% 
  filter(ctrl_perturbation == FALSE) %>% 
  bind_rows(etp_ctrl) %>% 
  arrange(desc(p_value))

# add pseudocount to Hi-C interaction frequencies
etp <- mutate(etp, hic_int_freq = hic_int_freq + 1)
```

***

## Chromatin features across ETP classes
Plot chromatin features of all `r sum(etp$significant, na.rm = TRUE)` hits across all classes of
perturbed target-gene pairs, including promoter and distal enhancer targeting perturbations.

```{r redefineClasses}
etp <- etp %>% 
  filter(pert_chr == gene_chr)
  
```


```{r prepareData1}
# features to include in plot
features <- c("log_2_fold_change", "enh_signal_count_DNase-seq", "enh_signal_avg_H3K27ac",
              "enh_signal_avg_H3K4me1", "enh_signal_avg_H3K4me3", "hic_int_freq")

# extract features for significant ETPs
etp_feat <- etp %>% 
  select(response_id, grna_target, class_manuscript, significant, dist_to_tss, likely_trans_effect, ctrl_perturbation,
         all_of(features)) %>% 
  pivot_longer(cols = all_of(features), names_to = "feature", values_to = "value") %>% 
  mutate(dist_to_tss = abs(dist_to_tss))

# reformat feature names for plot
etp_feat <- etp_feat %>% 
  mutate(feature = case_when(
    feature == "log_2_fold_change" ~ "Effect size",
    feature == "hic_int_freq" ~ "Hi-C 3D contact",
    TRUE ~ sub(".+_(.+)", "\\1", feature)
  ))

# convert classes to factors for correct ordering
class_levels <- c("promoter-proximal", "promoter-distal", "intragenic-proximal",
                  "intragenic-distal","intergenic-distal")
etp_feat <- etp_feat %>% 
  mutate(class_manuscript = factor(class_manuscript, levels = rev(class_levels)))

# extract positive hits only for plots
etp_feat_sig <- filter(etp_feat, significant == TRUE & ctrl_perturbation==FALSE)

# add effect size as separate column
etp_feat_sig <- etp_feat_sig %>% 
  filter(feature == "Effect size") %>% 
  select(response_id, grna_target, effect_size = value) %>%
  left_join(etp_feat_sig, ., by = c("response_id", "grna_target"))

# add group for coloring based on element location
etp_feat_sig <- mutate(etp_feat_sig, class_color = class_manuscript)
```

The median feature signals are computed to use as baseline reference in subsequent plots.
```{r calcAverageFeatureValues}
# calculate average feature values across all tested ETP
avg_feature_values <- etp_feat %>% 
  group_by(feature) %>% 
  summarize(mean = mean(value, na.rm = TRUE),
            median = median(value, na.rm = TRUE))

# calculate average feature values across all negative tested elements
positive_elements <- unique(pull(filter(etp_feat, significant == TRUE), grna_target))
avg_feature_values_negatives <- etp_feat %>% 
  filter(!grna_target %in% positive_elements) %>% 
  group_by(feature) %>% 
  summarize(mean = mean(value, na.rm = TRUE),
            median = median(value, na.rm = TRUE))
```

Plot chromatin feature value distributions of positive hits across the different ETP classes for
each assay.
```{r customPlotTheme}
custom_theme <- function(...) {
  theme_bw(...) %+replace% {
    theme(
      axis.title.y = element_blank(), panel.grid = element_blank(),
      strip.background = element_blank(),
      axis.title.x = element_text(size = 9, color = "black"),
      axis.text.x = element_text(size = 9, color = "black"),
      axis.text.y = element_text(size = 9, color = "black"),
      strip.text = element_text(size = 11, color = "black")
    )
  }
}
```

### Colored by element position
Make plot with classes colored by element location (promoter, intragenic, intergenic).
```{r plotChromClassColor}
# colors for each class
class_colors <- c(
  "promoter-proximal" = "#0263a3",        #9ad8fc
  "promoter-distal" = "#4299d4",  #6ebff5 
  "intragenic-proximal" ="#9ad8fc",#"#daa520",
  "intragenic-distal" = "#e57373", #f2c94c",
  "intergenic-distal"= "#c62828" # "#e57373"
)

true_enhancer_colors <- c(
  "TRUE" = "#0263a3",        #9ad8fc
  "FALSE" = "#c62828" # "#e57373"
)

# make effect size plot
es_dat <- filter(etp_feat_sig, feature == "Effect size")
es_plot <- ggplot(es_dat, aes(y = class_manuscript, x = value, color = true_enhancer)) +
  facet_wrap(~feature, nrow = 1, scales = "free_x") +
  geom_jitter(height = 0.2, pch = 16) +
  geom_boxplot(outlier.shape = NA, fill = NA, width = 0.5, color = "black") +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs(x = expression("log"[2] ~ "fold change"), color = "Element\nposition") +
  scale_color_manual(values = true_enhancer_colors) +
  custom_theme()

# make DNase-seq plot
dnase_dat <- filter(etp_feat_sig, feature == "DNase-seq")
dnase_plot <- ggplot(dnase_dat,
                     aes(y = class_manuscript, x = value, color = true_enhancer)) +
  facet_wrap(~feature, nrow = 1, scales = "free_x") +
  geom_jitter(height = 0.2, pch = 16) +
  geom_boxplot(outlier.shape = NA, fill = NA, width = 0.5, color = "black") +
  geom_vline(data = filter(avg_feature_values, feature == "DNase-seq"),
             aes(xintercept = median), linetype = "dashed") +
  labs(x = "Read depth normalized\nsignal", color = "Element\nposition") +
  scale_color_manual(values = true_enhancer_colors) +
  scale_x_log10() +
  custom_theme()

# make ChIP-seq plot
chip_assays <- c("H3K27ac", "H3K4me1", "H3K4me3")
chip_data <- filter(etp_feat_sig, feature %in% chip_assays)
chip_plot <- ggplot(chip_data,
                    aes(y = class_manuscript, x = value, color = true_enhancer)) +
  facet_wrap(~feature, nrow = 1, scales = "free_x") +
  geom_jitter(height = 0.2, pch = 16) +
  geom_boxplot(outlier.shape = NA, fill = NA, width = 0.5, color = "black") +
  geom_vline(data = filter(avg_feature_values, feature %in% chip_assays),
             aes(xintercept = median), linetype = "dashed") +  
  labs(x = "Fold change over control", color = "Element\nposition") +
  scale_x_log10() +
  scale_color_manual(values = true_enhancer_colors) +
  custom_theme()

# make Hi-C plot
hic_data <- filter(etp_feat_sig, feature == "Hi-C 3D contact")
hic_plot <- ggplot(hic_data,
                   aes(y = class_manuscript, x = value, color = true_enhancer)) +
  facet_wrap(~feature, nrow = 1, scales = "free_x") +
  geom_jitter(height = 0.2, pch = 16) +
  geom_boxplot(outlier.shape = NA, fill = NA, width = 0.5, color = "black") +
  labs(x = "Hi-C interaction\nfrequency + 1", color = "Element\nposition") +
  scale_x_log10() +
  scale_color_manual(values = true_enhancer_colors) +
  custom_theme()
```

```{r arrangeChromClassColorPlot1, fig.height=4, fig.width=12}
# extract legend from one plot
legend <- get_legend(es_plot)

# arrange all plots into one set
panels <- plot_grid(
  es_plot + theme(legend.position = "none"),
  dnase_plot + theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(),
                     legend.position = "none"),
  chip_plot + theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(),
                    legend.position = "none"),
  hic_plot + theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(),
                   legend.position = "none"),
  nrow = 1, align = "h", rel_widths = c(0.9, 0.5, 1.4, 0.5, 0.5), scale = 0.95
)

# combine with legend to assemble entire panel
plot_grid(panels, legend, rel_widths = c(1, 0.1))
```

```{r arrangeChromClassColorPlot2, fig.height=4, fig.width=10.25}
# extract legend from one plot
legend <- get_legend(es_plot)

# arrange all plots into one set
panels <- plot_grid(
  es_plot + theme(legend.position = "none"),
  dnase_plot + theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(),
                     legend.position = "none"),
  chip_plot + theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(),
                    legend.position = "none"),
  nrow = 1, align = "h", rel_widths = c(0.9, 0.5, 1.4, 0.5), scale = 0.95
)

# combine with legend to assemble entire panel
plot_grid(panels, legend, rel_widths = c(1, 0.1))
```

### Colored by distance
Make plot colored by distance to TSS.
```{r plotChromClassesDistColor}
# make effect size plot

#etp_feat_sig <- etp_feat_sig %>%
  #mutate(class_manuscript = case_when(
   # class_manuscript == "intragenic-distal" & likely_trans_effect == TRUE ~ "intragenic-distal-trans",
   # class_manuscript == "intragenic-distal" & likely_trans_effect == FALSE ~ "intragenic-distal-cis",
   # TRUE ~ class_manuscript
#  ))

es_dat <- filter(etp_feat_sig, feature == "Effect size")
es_plot <- ggplot(es_dat, aes(y = class_manuscript, x = value, color = log10(dist_to_tss + 1))) +
  facet_wrap(~feature, nrow = 1, scales = "free_x") +
  geom_jitter(height = 0.2, pch = 16) +
  geom_boxplot(outlier.shape = NA, fill = NA, width = 0.5, color = "black") +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs(x = expression("log"[2] ~ "fold change"), color = "Dist. TSS") +
  scale_colour_gradient(low = "#d10202", high = "#025fc9") +
  custom_theme()

# make DNase-seq plot
dnase_dat <- filter(etp_feat_sig, feature == "DNase-seq")
dnase_plot <- ggplot(dnase_dat,
                     aes(y = class_manuscript, x = value, color = log10(dist_to_tss + 1))) +
  facet_wrap(~feature, nrow = 1, scales = "free_x") +
  geom_jitter(height = 0.2, pch = 16) +
  geom_boxplot(outlier.shape = NA, fill = NA, width = 0.5, color = "black") +
  geom_vline(data = filter(avg_feature_values, feature == "DNase-seq"),
             aes(xintercept = median), linetype = "dashed") +
  labs(x = "Read depth normalized\nsignal", color = "Dist. TSS") +
  scale_colour_gradient(low = "#d10202", high = "#025fc9") +
  scale_x_log10() +
  custom_theme()

# make ChIP-seq plot
chip_assays <- c("H3K27ac", "H3K4me1", "H3K4me3")
chip_data <- filter(etp_feat_sig, feature %in% chip_assays)
chip_plot <- ggplot(chip_data,
                    aes(y = class_manuscript, x = value, color = log10(dist_to_tss + 1))) +
  facet_wrap(~feature, nrow = 1, scales = "free_x") +
  geom_jitter(height = 0.2, pch = 16) +
  geom_boxplot(outlier.shape = NA, fill = NA, width = 0.5, color = "black") +
  geom_vline(data = filter(avg_feature_values, feature %in% chip_assays),
             aes(xintercept = median), linetype = "dashed") +  
  labs(x = "Fold change over control", color = "Dist. TSS") +
  scale_x_log10() +
  scale_colour_gradient(low = "#d10202", high = "#025fc9") +
  custom_theme()

# make Hi-C plot
hic_data <- filter(etp_feat_sig, feature == "Hi-C 3D contact")
hic_plot <- ggplot(hic_data,
                   aes(y = class_manuscript, x = value, color = log10(dist_to_tss + 1))) +
  facet_wrap(~feature, nrow = 1, scales = "free_x") +
  geom_jitter(height = 0.2, pch = 16) +
  geom_boxplot(outlier.shape = NA, fill = NA, width = 0.5, color = "black") +
  labs(x = "Hi-C interaction\nfrequency + 1", color = "Dist. TSS") +
  scale_x_log10() +
  scale_colour_gradient(low = "#d10202", high = "#025fc9") +  
  custom_theme()
```

```{r arrangeChromClassesDistColorPlot, fig.height=4, fig.width=12}
# extract legend from one plot
legend <- get_legend(es_plot)

# arrange all plots into one set
panels <- plot_grid(
  es_plot + theme(legend.position = "none"),
  dnase_plot + theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(),
                     legend.position = "none"),
  chip_plot + theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(),
                    legend.position = "none"),
  hic_plot + theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(),
                   legend.position = "none"),
  nrow = 1, align = "h", rel_widths = c(0.9, 0.5, 1.4, 0.5, 0.5), scale = 0.95
)

# combine with legend to assemble entire panel
plot_grid(panels, legend, rel_widths = c(1, 0.1))
```

### Colored by effect size
Make plot colored by CRISPRi effect sizes.
```{r plotChromClassesESColor}
# make effect size plot
es_dat <- filter(etp_feat_sig, feature == "Effect size")
es_plot <- ggplot(es_dat, aes(y = class_manuscript, x = value, color = effect_size)) +
  facet_wrap(~feature, nrow = 1, scales = "free_x") +
  geom_jitter(height = 0.2, pch = 16) +
  geom_boxplot(outlier.shape = NA, fill = NA, width = 0.5, color = "black") +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs(x = expression("log"[2] ~ "fold change"), color = "Effect size\n(log2 FC)") +
  scale_colour_gradientn(colors = c("red", "red3", "purple3", "blue3", "blue")) +
  custom_theme()

# make DNase-seq plot
dnase_dat <- filter(etp_feat_sig, feature == "DNase-seq")
dnase_plot <- ggplot(dnase_dat,
                     aes(y = class_manuscript, x = value, color = effect_size)) +
  facet_wrap(~feature, nrow = 1, scales = "free_x") +
  geom_jitter(height = 0.2, pch = 16) +
  geom_boxplot(outlier.shape = NA, fill = NA, width = 0.5, color = "black") +
  geom_vline(data = filter(avg_feature_values, feature == "DNase-seq"),
             aes(xintercept = median), linetype = "dashed") +
  labs(x = "Read depth normalized\nsignal", color = "Effect size\n(log2 FC)") +
  scale_colour_gradientn(colors = c("red", "red3", "purple3", "blue3", "blue")) +
  scale_x_log10() +
  custom_theme()

# make ChIP-seq plot
chip_assays <- c("H3K27ac", "H3K4me1", "H3K4me3")
chip_data <- filter(etp_feat_sig, feature %in% chip_assays)
chip_plot <- ggplot(chip_data,
                    aes(y = class_manuscript, x = value, color = effect_size)) +
  facet_wrap(~feature, nrow = 1, scales = "free_x") +
  geom_jitter(height = 0.2, pch = 16) +
  geom_boxplot(outlier.shape = NA, fill = NA, width = 0.5, color = "black") +
  geom_vline(data = filter(avg_feature_values, feature %in% chip_assays),
             aes(xintercept = median), linetype = "dashed") +  
  labs(x = "Fold change over control", color = "Effect size\n(log2 FC)") +
  scale_x_log10() +
  scale_colour_gradientn(colors = c("red", "red3", "purple3", "blue3", "blue")) +
  custom_theme()

# make Hi-C plot
hic_data <- filter(hic_data, feature == "Hi-C 3D contact")
hic_plot <- ggplot(hic_data,
                   aes(y = class_manuscript, x = value, color = effect_size)) +
  facet_wrap(~feature, nrow = 1, scales = "free_x") +
  geom_jitter(height = 0.2, pch = 16) +
  geom_boxplot(outlier.shape = NA, fill = NA, width = 0.5, color = "black") +
  labs(x = "Hi-C interaction\nfrequency + 1", color = "Effect size\n(log2 FC)") +
  scale_x_log10() +
  scale_colour_gradientn(colors = c("red", "red3", "purple3", "blue3", "blue")) +
  custom_theme()
```

```{r arrangeChromClassesESColorPlot, fig.height=4, fig.width=12}
# extract legend from one plot
legend <- get_legend(es_plot)

# arrange all plots into one set
panels <- plot_grid(
  es_plot + theme(legend.position = "none"),
  dnase_plot + theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(),
                     legend.position = "none"),
  chip_plot + theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(),
                    legend.position = "none"),
  hic_plot + theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(),
                   legend.position = "none"),
  nrow = 1, align = "h", rel_widths = c(0.9, 0.5, 1.4, 0.5, 0.5), scale = 0.95
)

# combine with legend to assemble entire panel
plot_grid(panels, legend, rel_widths = c(1, 0.1))
```

***

## Intragenic enhancers
```{r, fig.width=8, fig.height=3}
# extract chromatin features for intragenic distal enhancers
intragenic_enh <- etp_feat_sig %>% 
  filter(class_manuscript == "intragenic-distal") %>% 
  mutate(feature = fct_relevel(feature, "Effect size"))

# extract results for enhancers within their host gene
intragenic_proximal_pairs <- filter(etp, class_manuscript == "intragenic-proximal")

# get the distance to host gene tss for each perturbation
dist_to_host_tss <- etp %>% 
  filter(class_manuscript == "intragenic-proximal") %>% 
  group_by(grna_target) %>% 
  summarize(dist_to_host_tss = min(abs(dist_to_tss)))

# add distance to host gene to intragenic-distal hits
intragenic_enh <- left_join(intragenic_enh, dist_to_host_tss, by = "grna_target")

# get intragenic perturbations with effects on their host genes
host_hits <- intragenic_proximal_pairs %>% 
  filter(significant == TRUE) %>% 
  pull(grna_target) %>% 
  unique()

# add information to intragenic enhancers
intragenic_enh <- mutate(intragenic_enh, host_hit = grna_target %in% host_hits)
```

Look at position of intragenic enhancers within host genes and overlap with hits on host genes.
```{r}
# select effect size feature only for overall property plots
intragenic_enh_es <- filter(intragenic_enh, feature == "Effect size")

# plot proportion of host hits
host_hits_counts <- count(intragenic_enh_es, likely_trans_effect, host_hit, name = "hits")
plot_hist_hits <- ggplot(host_hits_counts, aes(x = likely_trans_effect, y = hits, fill = host_hit)) +
  geom_bar(stat = "identity") +
  labs(title = "Effects on host gene", y = "Hits", x = "Gene-gene interaction",
       fill = "Host gene hit") +
  scale_fill_manual(values = c("TRUE" = "goldenrod1", "FALSE" = "turquoise3")) +
  theme_bw() +
  theme(panel.grid = element_blank())

# plot distance to host gene for intragenic enhancers
plot_host_dist <- intragenic_enh_es %>% 
  filter(feature == "Effect size") %>% 
  ggplot(., aes(x = likely_trans_effect, y = dist_to_host_tss / 1000, color = likely_trans_effect)) +
    geom_jitter(width = 0.2) +
    geom_boxplot(fill = NA, color = "black", outlier.shape = NA, width = 0.5) +
    labs(y = "Distance (kb)", x = "Gene-gene interaction", title = "Distance to host TSS") +
    scale_color_manual(values = c("TRUE" = "red3", "FALSE" = "steelblue")) +
    theme_bw() +
    theme(panel.grid = element_blank(), legend.position = "none")
```

```{r, fig.width=7, fig.height=3}
# arrange intragenic enhancer properties into one figure
plot_grid(plot_hist_hits, plot_host_dist, nrow = 1, scale = 0.95, rel_widths = c(1, 0.75))
```

```{r, fig.height=3, fig.width=9}
# set levels for features in average values table
avg_feature_values <- mutate(avg_feature_values, feature = fct_relevel(feature, "Effect size"))

# plot chromatin features between likely trans effects and others
ggplot(intragenic_enh, aes(x = likely_trans_effect, y = value, color = likely_trans_effect)) +
  facet_wrap(~feature, nrow = 1, scales = "free") +
  geom_jitter(width = 0.2) +
  geom_boxplot(fill = NA, color = "black", outlier.shape = NA, width = 0.5) +
  geom_hline(data = avg_feature_values, aes(yintercept = median), linetype = "dashed") +  
  labs(x = "Likely trans-acting interaction") +
  scale_color_manual(values = c("TRUE" = "red3", "FALSE" = "steelblue")) +
  theme_bw() +
  theme(panel.grid = element_blank(), strip.background = element_blank(), legend.position = "none")
```

```{r}
# filter out any intragenic-distal ETPs that are likely trans-effects
etp_filt <- filter(etp, !(class_manuscript == "intragenic-distal" & likely_trans_effect == TRUE))
```

***

## Distal enhancers
Plot chromatin features of distal enhancers as function of distance to TSS.
```{r prepareData2}
# recalculate distance to TSS categories
etp <- etp %>% 
  mutate(distToTSS_bin = case_when(
    class_manuscript == "promoter-proximal" ~ "TSS",
    class_manuscript %in% c("intergenic-distal", "intragenic-distal", "intragenic-proximal") & (abs(dist_to_tss) > 1500 & abs(dist_to_tss) <= 1e4) ~ "1500-10k",
    class_manuscript %in% c("intergenic-distal", "intragenic-distal", "intragenic-proximal") & (abs(dist_to_tss) > 1e4 & abs(dist_to_tss) <= 1e5) ~ "10k-100k",
    class_manuscript %in% c("intergenic-distal", "intragenic-distal", "intragenic-proximal") & (abs(dist_to_tss) > 1e5 & abs(dist_to_tss) <= 1e6) ~ "100k-1Mb",
    class_manuscript %in% c("intergenic-distal", "intragenic-distal", "intragenic-proximal") ~ ">1Mb",
    TRUE ~ NA_character_
  ))

# make distance to TSS bin an ordered factor
distToTSS_bin_levels <- c("TSS", "1500-10k", "10k-100k", "100k-1Mb", ">1Mb")
etp <- mutate(etp, distToTSS_bin = fct_relevel(distToTSS_bin, distToTSS_bin_levels))

# filter for specified E-G pairs to include in analysis
enh_etp <- etp %>%
  filter(!is.na(significant)) %>% 
  filter(class_manuscript %in% c("intergenic-distal","intragenic-distal", "intragenic-proximal", "promoter-proximal"))

# chromatin assays to include in plot
chrom_assays <- c("enh_signal_count_DNase-seq", "enh_signal_avg_H3K27ac", "enh_signal_avg_H3K4me1",
                  "enh_signal_avg_H3K4me3")

# select any elements that were detected as positives and categorize by distance
all_bins <- c("hit TSS", "hit 1500-10k", "hit 10k-100k", "hit 100k-1Mb", "hit >1Mb", "NS")
positives <- enh_etp %>% 
  filter(significant == TRUE) %>% 
  mutate(category = paste("hit", distToTSS_bin)) %>%
  mutate(category = factor(category, levels = all_bins, ordered = TRUE) )

# select unique enhancers and assays to plot
positives <- positives %>% 
  select(grna_target, significant, category, all_of(chrom_assays)) %>% 
  distinct()

# if an enhancer is counted in multiple categories, assign it to the closest distance to TSS bin
positives <- positives %>% 
  group_by(grna_target) %>% 
  slice_min(category, n = 1)

# filter for enhancers that are not linked to any gene and select columns for plot
negatives <- enh_etp %>% 
  filter(significant == FALSE, !grna_target %in% positives$grna_target) %>% 
  mutate(category = factor("NS", levels = all_bins, ordered = TRUE)) %>% 
  select(grna_target, significant, category, all_of(chrom_assays)) %>% 
  distinct()

# combine back into one table and convert to long format for plots
plot_etp <- bind_rows(positives, negatives) %>% 
  pivot_longer(cols = all_of(chrom_assays),
               names_to = "column_name", values_to = "signal") %>% 
  mutate(column_name = sub("enh_signal_", "", column_name)) %>% 
  separate(column_name, into = c("metric", "assay"), sep = "_")
```

```{r makeChromDistPlot}
# perform statistical tests (kruskal-wallis) between categories
kw_tests <- plot_etp %>% 
  ungroup() %>% 
  compare_means(formula = signal ~ category, method = "kruskal.test", data = ., group.by = "assay") %>% 
  mutate(p.adj.plot = if_else(p.adj < 0.001, true = "< 0.001",
                              false = paste("=", round(p.adj, digits = 2))))

# get sample sizes per category group
sample_sizes <- plot_etp %>% 
  ungroup() %>% 
  select(grna_target, category) %>% 
  distinct() %>% 
  arrange(category) %>% 
  count(category, .drop = FALSE)

# set new labels (factor levels) for each category, which include sample size per category
levels <- paste0(sample_sizes$category, " (n=", sample_sizes$n, ")")
levels(plot_etp$category) <- levels

# add mock group to KW test table
kw_tests <- mutate(kw_tests, group1 = levels[[2]], group2 = levels[[3]])

# colors for enhancer categories in figures
category_colors <- structure(c("#daa520", "#0263a3", "#4299d4", "#6ebff5", "#9ad8fc", "#a9a9a9"),
                             names = levels)

# plot DNase-seq accessibility
p1 <- ggplot(filter(plot_etp, assay == "DNase-seq"),
             aes(x = category, y = signal + 1, fill = category)) +
  facet_wrap(~ assay, nrow = 1) +
  geom_boxplot() +
  stat_pvalue_manual(data = filter(kw_tests, assay == "DNase-seq"), y.position = 15000, size = 3,
                     label = "P adj. {p.adj.plot}", remove.bracket = TRUE, inherit.aes = FALSE) +
  labs(y = paste0("Read-depth normalized", "\n", "signal")) +
  scale_fill_manual(values = category_colors) +
  scale_y_log10(limits = c(1, 20000)) +
  theme_bw() +
  theme(legend.position = "none", axis.title.x = element_blank(), panel.grid = element_blank(),
        axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))

# plot ChIP-seq signals
p2 <- ggplot(filter(plot_etp, assay %in% c("H3K27ac", "H3K4me1", "H3K4me3")),
             aes(x = category, y = signal + 0.01, fill = category)) +
  facet_wrap(~ assay, nrow = 1) +
  geom_boxplot() +
  stat_pvalue_manual(data = filter(kw_tests, assay %in% c("H3K27ac", "H3K4me1", "H3K4me3")),
                     y.position = 120, size = 3, label = "P adj. {p.adj.plot}",
                     remove.bracket = TRUE, inherit.aes = FALSE) +
  labs(y = "FC over control") +
  scale_fill_manual(values = category_colors) +
  scale_y_log10(limits = c( 0.01, 150)) +
  theme_bw() +
  theme(legend.position = "none", axis.title.x = element_blank(), panel.grid = element_blank(),
        axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
```

```{r arrangeChromDistPlot, fig.height=3, fig.width=7}
# assemble figure panel and save to pdf
plot_grid(p1, p2, nrow = 1, rel_widths = c(0.5, 1))
```

```{r ChromDistPWTests}
# perform individual pairwise tests
pw_tests <- compare_means(signal ~ category, group.by = "assay", data = ungroup(plot_etp),
                          method = "wilcox.test", p.adjust.method = "holm")

# split category labels in to labels and number of enhancers
pw_tests <- pw_tests %>% 
  separate(group1, into = c("group1", "n_group1"), sep = " \\(n=") %>% 
  separate(group2, into = c("group2", "n_group2"), sep = " \\(n=") %>% 
  mutate(n_group1 = sub("\\)", "", n_group1), n_group2 = sub("\\)", "", n_group2)) %>% 
  select(-.y.)

# save to text file
write_csv(pw_tests, file = "results/tables/chromatin_features_distal_pairwise_tests.csv")
```

***

## 3D interaction frequency
Plot Hi-C 3D interaction frequencies of distal interactions as function of distance to TSS.
```{r prepareData3}
# add significance label for Hi-C plot
enh_etp <- enh_etp %>% 
  mutate(sig_label = if_else(significant, true = "Sig.", false = "NS")) %>% 
  add_count(sig_label) %>% 
  mutate(sig_label = paste0(sig_label, " (n=", n, ")")) %>% 
  mutate(sig_label = fct_reorder(sig_label, n)) %>% 
  select(-n)

# create new distance to TSS bin levels which include number of pairs
distToTSS_levels <- enh_etp %>% 
  group_by(distToTSS_bin) %>% 
  summarize(sig = sum(significant == TRUE),
            nonsig = sum(significant == FALSE)) %>% 
  mutate(distToTSS_bin = as.character(distToTSS_bin)) %>% 
  mutate(distToTSS_label = paste0(distToTSS_bin, "\nSig.=", sig, "\nNS=", nonsig)) %>% 
  select(distToTSS_label, distToTSS_bin) %>% 
  deframe()

# change distance to TSS labels in ETP table
enh_etp <- mutate(enh_etp, distToTSS_bin = fct_recode(distToTSS_bin, !!!distToTSS_levels))

# set colors for significance labels which include number of pairs
sig_cols <- structure(c("steelblue", "darkgray"), names = levels(enh_etp$sig_label))

# set NA interaction frequency to 0
enh_etp <- replace_na(enh_etp, replace = list(hic_int_freq = 0))

# calculate distance to TSS in bin increments
enh_etp <- enh_etp %>% 
  mutate(hic_bin_dist = abs(dist_to_tss) %/% snakemake@params$hic_res * snakemake@params$hic_res)
```

```{r plotHicDistBins, fig.height=3.5, fig.width=6}
# plot interaction frequencies per distance to TSS bin
ggplot(enh_etp, aes(x = distToTSS_bin, y = hic_int_freq + 1, fill = sig_label)) +
  geom_boxplot() +
  labs(y = "Hi-C interaction frequency + 1", x = "Distance to TSS", fill = "Significance") +
  scale_fill_manual(values = sig_cols) +
  scale_y_log10(limits = c(1, 1000)) +
  theme_bw()
```

```{r plotHicDist, fig.height=3.5, fig.width=6.5}
# only look at results within 1Mb (typical enhancer range)
enh_etp_filt <- enh_etp %>% 
  filter(abs(dist_to_tss) <= 1e6) %>% 
  mutate(sig_label = if_else(significant, true = "Sig.", false = "NS")) %>% 
  add_count(sig_label) %>% 
  mutate(sig_label = paste0(sig_label, " (n=", n, ")")) %>% 
  mutate(sig_label = fct_reorder(sig_label, n)) %>% 
  select(-n)

# make simple 2 group boxplot
p1 <- ggplot(enh_etp_filt, aes(x = sig_label, y = hic_int_freq + 1, fill = sig_label)) +
  geom_boxplot() +
  labs(y = "Hi-C interaction frequency + 1", x = "") +
  scale_fill_manual(values = c("steelblue", "gray66")) +
  scale_y_log10(limits = c(1, 1000)) +
  theme_bw() +
  theme(legend.position = "none", axis.text.x = element_text(angle = 45, hjust = 1))

# plot interaction frequencies as function of distance to TSS
p2 <- ggplot(enh_etp_filt, aes(x = abs(dist_to_tss) / 1000, y = hic_int_freq + 1, color = sig_label)) +
  geom_point(data = filter(enh_etp_filt, significant == FALSE)) +
  geom_point(data = filter(enh_etp_filt, significant == TRUE)) +
  geom_smooth(data = filter(enh_etp_filt, significant == FALSE), se = FALSE, color = "gray40") +
  geom_smooth(data = filter(enh_etp_filt, significant == TRUE), se = FALSE, color = "steelblue") +
  labs(y = "Hi-C interaction frequency + 1", x = "Distance to TSS (kb)", color = "Significance") +
  scale_color_manual(values = c("gray66", "steelblue")) +
  scale_x_log10(limits = c(1, NA)) +
  scale_y_log10(limits = c(1, 1000)) +
  theme_bw() +
  theme(axis.title.y = element_blank(),
        axis.text.y = element_blank())

# arrange into one figure and save to pdf file
plot_grid(p1, p2, nrow = 1, rel_widths = c(0.27, 0.73), align = "h")
```

```{r plotHicDistBinsPWTests}
# perform individual pairwise tests
pw_tests <- compare_means(hic_int_freq ~ significant, data = enh_etp, group.by = "distToTSS_bin",
                          method = "wilcox.test", p.adjust.method = "holm")

# split category labels in to labels and number of enhancers
pw_tests <- pw_tests %>% 
  select(-c(.y., group1, group2)) %>% 
  separate(distToTSS_bin, into = c("distToTSS_bin", "group1", "n_group1", "group2", "n_group2"),
           sep = "\n|=")

# save to text file
write_csv(pw_tests, file = "results/tables/hic_interaction_frequency_distBins_pairwise_tests.csv")
```

***

```{r}
discovery_P <-read_tsv("/g/steinmetz/moonen/Rstudio/significant_discovery_pairs.tsv")
positive_C <-read_tsv("/g/steinmetz/moonen/Rstudio/positive_controls.tsv")

pos_neg_genes <- positive_C %>%
  filter(
    pass_qc == TRUE,
    !is.na(log_2_fold_change),
    log_2_fold_change < 0,
    !is.na(empirical_pvalue),
    empirical_pvalue < 0.05
  ) %>%
  pull(gene_symbol) %>%
  unique()




# Unique CRE–gene–class pairs from es_dat
es_pairs <- es_dat %>%
  distinct(
    es_grna_target   = grna_target,
    es_response_id   = response_id,
    class_manuscript,
    effect_size
  )

# edges1: CRE -> primary gene
edges1 <- es_pairs %>%
  transmute(
    from       = es_grna_target,
    to         = es_response_id,
    log2fc     = effect_size,  # or abs(effect_size) later for width
    edge_type  = "CRE_to_gene",
    edge_color = class_colors[class_manuscript]
  )

# join es_pairs (primary gene) to discovery_P (grna_target)
matches <- es_pairs %>%
  inner_join(
    discovery_P %>%
      select(grna_target, gene_symbol, log_2_fold_change),
    by = c("es_response_id" = "grna_target")
  )

# edges2: primary gene -> downstream gene_symbol
edges2 <- matches %>%
  transmute(
    from       = es_response_id,
    to         = gene_symbol,
    log2fc     = log_2_fold_change,
    edge_type  = "gene_to_gene",
    edge_color = NA_character_
  )

map_dp <- discovery_P %>%
  select(grna_target, gene_symbol, log_2_fold_change)

frontier <- matches %>%
  select(from = gene_symbol) %>%   # start from downstream genes
  distinct()

all_follow_edges <- tibble(
  from      = character(),
  to        = character(),
  log2fc    = numeric(),
  edge_type = character()
)

seen <- frontier$from

repeat {
  step <- frontier %>%
    inner_join(map_dp, by = c("from" = "grna_target")) %>%
    transmute(
      from,
      to        = gene_symbol,
      log2fc    = log_2_fold_change,
      edge_type = "dp_chain"
    ) %>%
    distinct()

  if (nrow(step) == 0) break

  all_follow_edges <- bind_rows(all_follow_edges, step)

  new_nodes <- setdiff(step$to, seen)
  if (length(new_nodes) == 0) break

  frontier <- tibble(from = new_nodes)
  seen <- union(seen, new_nodes)
}

edges3_recursive <- all_follow_edges %>%
  mutate(edge_color = NA_character_)

edge_list <- bind_rows(edges1, edges2, edges3_recursive) %>%
  group_by(from, to) %>%
  summarise(
    log2fc     = first(log2fc),                          # or mean(log2fc), max(abs(log2fc))
    edge_type  = paste(unique(edge_type), collapse = ";"),  # keep track of origins if useful
    edge_color = first(na.omit(edge_color)),
    .groups = "drop"
  )

nodes <- tibble(name = unique(c(edge_list$from, edge_list$to))) %>%
  mutate(
    type = case_when(
      name %in% es_pairs$es_grna_target ~ "CRE",
      TRUE                              ~ "gene"
    )
  )

# Identify self-targeting genes in discovery_P
self_target_genes <- discovery_P %>%
  filter(grna_target == gene_symbol) %>%
  pull(gene_symbol) %>%
  unique()

# Update node types
nodes <- tibble(name = unique(c(edge_list$from, edge_list$to))) %>%
  mutate(
    type = case_when(
      name %in% es_pairs$es_grna_target ~ "CRE",          # CREs
      name %in% pos_neg_genes           ~ "pos_neg_gene", # positive controls: sig & negative
      TRUE                              ~ "gene"          # all other genes
    )
  )

g <- graph_from_data_frame(edge_list, directed = TRUE, vertices = nodes)

# Edge colors: class_colors for CRE->gene, grey otherwise
E(g)$color <- ifelse(!is.na(E(g)$edge_color), E(g)$edge_color, "grey60")

# Edge widths: scale |log2fc|, but note edges1 now use effect_size
E(g)$width <- rescale(abs(E(g)$log2fc), to = c(1, 6))

color_map <- c(
  CRE          = "red",
  gene         = "dodgerblue3",
  pos_neg_gene = "forestgreen"
)



V(g)$shape <- "circle"
V(g)$color <- color_map[V(g)$type]
V(g)$size  <- 2

plot(
  g,
  vertex.label  = NA,
  edge.arrow.size = 0.01,
  layout = layout_with_fr(g)
)
```

filter for only those networks that have at least 1 CRE, and for those genes that have an effect on itself (control genes), but keep non-control genes when its an end node. 
```{r}
## 1. Split edges into CRE→gene vs gene→gene -------------------------------

# assumes you set edge_type when building the graph:
#   "CRE_to_gene"  for CRE -> primary gene
#   "gene_to_gene" and "dp_chain" for gene -> gene edges

is_cre_edge <- E(g)$edge_type == "CRE_to_gene"

cre_gene_edges  <- E(g)[is_cre_edge]         # keep ALL of these
non_cre_edges   <- E(g)[!is_cre_edge]        # gene→gene edges (and similar)


## 2. From gene→gene edges, keep only those where the source is green -----

gene_edges_keep <- non_cre_edges[0]  # empty by default

if (length(non_cre_edges) > 0) {
  # ends() gives from/to vertex indices if names = FALSE
  non_cre_ends <- ends(g, non_cre_edges, names = FALSE)
  from_idx     <- non_cre_ends[, 1]

  # assume green genes were tagged as type == "pos_neg_gene"
  is_from_green <- V(g)$type[from_idx] == "pos_neg_gene"

  gene_edges_keep <- non_cre_edges[is_from_green]
}

# all edges we want to keep
edges_keep <- c(cre_gene_edges, gene_edges_keep)


## 3. Build subgraph from these edges, dropping isolated nodes ------------

g_edges <- subgraph.edges(g, e = edges_keep, delete.vertices = TRUE)


## 4. Keep only components with at least one CRE AND at least one gene ----

comp <- components(g_edges)

comp_has_cre <- tapply(V(g_edges)$type == "CRE",
                       comp$membership, any)

comp_has_gene <- tapply(V(g_edges)$type %in% c("gene", "pos_neg_gene"),
                        comp$membership, any)

comps_to_keep <- which(comp_has_cre & comp_has_gene)

verts_keep <- V(g_edges)[comp$membership %in% comps_to_keep]

g_final <- induced_subgraph(g_edges, vids = verts_keep)


## 5. Plot -----------------------------------------------------------------

# Kamada-Kawai – often the most balanced
coords <- layout_with_kk(g_final)

# Graphopt – good for large networks
coords <- layout_with_graphopt(g_final)



# DrL – scalable for big graphs, hierarchical feel
coords <- layout_with_drl(g_final)

plot(
  g_final,
  layout            = coords,
  vertex.label      = NA,
  vertex.size       = V(g_final)$size,
  vertex.color      = V(g_final)$color,
  vertex.frame.color = NA,
  edge.color        = E(g_final)$color,
  edge.width        = E(g_final)$width,
  edge.arrow.size   = 0.01
)
```
keep only networks that include both CRE-gene and gene-gene connections. 
```{r}
## 1. Identify gene-to-gene edges (both ends are genes) on g_final ----
is_gene_gene_edge <- function(g) {
  ends_mat  <- ends(g, E(g), names = FALSE)
  from_type <- V(g)$type[ends_mat[,1]]
  to_type   <- V(g)$type[ends_mat[,2]]
  from_type %in% c("gene", "pos_neg_gene") &
    to_type %in% c("gene", "pos_neg_gene")
}

E(g_final)$gene_gene <- is_gene_gene_edge(g_final)

## 2. Filter to components that contain at least one gene→gene edge ----
comp <- components(g_final)

# map each edge to the component of its source vertex
edge_comp <- comp$membership[ ends(g_final, E(g_final), names = FALSE)[,1] ]

comp_has_gene_gene <- tapply(E(g_final)$gene_gene, edge_comp, any)

comps_keep  <- as.numeric(names(comp_has_gene_gene)[comp_has_gene_gene])
verts_keep  <- V(g_final)[ comp$membership %in% comps_keep ]

# this is now your cleaned graph
g_clean <- induced_subgraph(g_final, vids = verts_keep)

## 3. Centrality & core genes (on g_clean!) ----
# centrality
V(g_clean)$degree      <- degree(g_clean, mode = "all")
V(g_clean)$betweenness <- betweenness(g_clean, directed = TRUE, normalized = TRUE)
V(g_clean)$eigen       <- evcent(g_clean, directed = TRUE)$vector

# genes only (exclude CREs)
gene_nodes <- V(g_clean)[V(g_clean)$type %in% c("gene", "pos_neg_gene")]

# robust top 10% by degree: pick an explicit number of nodes
n_genes   <- length(gene_nodes)
n_core    <- max(1, ceiling(0.10 * n_genes))  # at least 1
ord_deg   <- order(gene_nodes$degree, decreasing = TRUE, na.last = NA)
core_genes <- gene_nodes[ ord_deg[seq_len(n_core)] ]

## 4. Labels only for core genes ----
V(g_clean)$label <- NA                     # clear all labels
V(g_clean)$label[core_genes] <- V(g_clean)$name[core_genes]  # label only cores
## 5. Plot ----
coords <- layout_with_fr(g_clean, niter = 3000)

plot(
  g_clean,
  layout           = coords,
  vertex.label     = V(g_clean)$label,
  vertex.label.cex = 0.8,
  vertex.size      = V(g_clean)$size,
  vertex.color     = V(g_clean)$color,
  vertex.frame.color = NA,
  edge.color       = E(g_clean)$color,
  edge.width       = E(g_clean)$width,
  edge.arrow.size  = 0.01
)
```
