---
title: "Enhancer-Promoter Network Analysis"
author: "Dewi Moonen"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
editor_options: 
  chunk_output_type: console
---


```{r}
# set output html chunk options
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```


```{r attachPackages}
library(tidyverse)
library(cowplot)
library(ggpubr)
library(igraph)
library(dplyr)

select <- dplyr::select
rename <- dplyr::rename
count <- dplyr::count
slice <- dplyr::slice
map <- purrr::map
mutate <- dplyr::mutate
filter<-dplyr::filter
first <-dplyr::first
components <-igraph::components
degree <-igraph::degree
simplify <-igraph::simplify
source("/g/steinmetz/moonen/Rstudio/theme_manuscript_default.R")

```

load data
```{r}
#etp <- read_csv("/g/steinmetz/project/otar_2063/enhancer_screen/interpretation/chromatin_analyses/results/results_df_with_promoterC_annotated_allFeatures_transHits.csv")
#etp <- read_csv("/g/steinmetz/project/otar_2063/enhancer_screen/interpretation/chromatin_analyses/results/results_df_with_promoterC_annotated_allFeatures.csv")
#etp<-read_csv("/g/steinmetz/project/otar_2063/enhancer_screen/interpretation/chromatin_analyses/results/share/cre_perturbation_screen_results.csv")
etp<-read_csv("/g/steinmetz/moonen/Rstudio/cre_perturbation_screen_results_traits.csv")

#discovery_P <-read_tsv("/g/steinmetz/moonen/Rstudio/significant_discovery_pairs.tsv")
#discovery_P <-read_tsv("/g/steinmetz/project/otar_2063/promoter_screen/interpretation/diff_expr_analyses/process_trans_analysis/sceptre_pvalues/discovery_pairs_significant_fdr001.tsv.gz")

#discovery_P <-read_tsv("/g/steinmetz/project/otar_2063/promoter_screen/interpretation/diff_expr_analyses/process_trans_analysis/gene_level_empirical_pvalues/upsample_nt_pvals/discovery_pairs_empirical_pvalues.tsv.gz")
#discovery_P <-read_tsv("/g/steinmetz/project/otar_2063/promoter_screen/interpretation/diff_expr_analyses/process_trans_analysis/gene_level_empirical_pvalues/significant_trans_pairs_per_gene_fdr.tsv.gz")



#positive_C <-read_tsv("/g/steinmetz/moonen/Rstudio/positive_controls.tsv")
#positive_C <-read_tsv("/g/steinmetz/project/otar_2063/promoter_screen/interpretation/diff_expr_analyses/process_trans_analysis/sceptre_pvalues/positive_control_results.tsv.gz")

#positive_C <-read_tsv("/g/steinmetz/project/otar_2063/promoter_screen/interpretation/diff_expr_analyses/process_trans_analysis/gene_level_empirical_pvalues/upsample_nt_pvals/positive_control_pairs_empirical_pvalues.tsv.gz")
#positive_C <-read_tsv("/g/steinmetz/project/otar_2063/promoter_screen/interpretation/diff_expr_analyses/process_trans_analysis/gene_level_empirical_pvalues/positive_control_pairs_empirical_pvalues.tsv.gz")

#filter discovery list to get cleaner network
#discovery_P <- discovery_P%>% filter(pval_adj_per_gene<=0.1, abs(log_2_fold_change) >= 0.1)

#discovery_P_full  <-read_tsv("/g/steinmetz/project/otar_2063/promoter_screen/interpretation/diff_expr_analyses/process_trans_analysis/sceptre_pvalues/discovery_pairs_significant_fdr001.tsv.gz" )
#positive_C_full  <-read_tsv("/g/steinmetz/project/otar_2063/promoter_screen/interpretation/diff_expr_analyses/process_trans_analysis/sceptre_pvalues/positive_control_results.tsv.gz" )

#discovery from Stefan
#folder_path <- "/g/stegle/schrod/code/TCell/results_pseudobulk/DEG_bulk_1_and_2_guides_15_nt"
#folder_path <- "/g/stegle/schrod/code/TCell/results/DEG"

folder_path <- "/g/stegle/schrod/code/TCell/pseudobulk_deseq2_chunks50_min3bulks_ribosomal"
discovery_P <- list.files(path = folder_path, pattern = "\\.csv$", full.names = TRUE) %>%
  map_df(read_csv)

discovery_P <- discovery_P %>% mutate(grna_target = contrast, gene_symbol = variable, log_2_fold_change = log_fc, pvals_adj = adj_p_value) %>% filter(pvals_adj <=0.1 , abs(log_2_fold_change) >= 0.1)

positive_C <- discovery_P%>%
  filter(gene_symbol == grna_target, 
    !is.na(log_2_fold_change),
    log_2_fold_change < 0) %>%
  pull(gene_symbol) %>%
  unique()

pos_neg_genes <- positive_C
```

create ETP list to use 
```{r}
# for control perturbations which were in multiple panels, select the most significant one
etp_ctrl <- etp %>% 
  filter(ctrl_perturbation == TRUE) %>% 
  group_by(response_id, grna_target) %>% 
  slice_min(p_value, n = 1, with_ties = FALSE)

# add back to results table with discovery pairs, filter for specific disease. 
etp <- etp %>% 
 # filter(str_detect(trait, "Inflammatory_bowel_disease")) %>%
  filter(ctrl_perturbation == FALSE) %>% 
  bind_rows(etp_ctrl) %>% 
  arrange(desc(p_value))

etp <- etp %>% 
  filter(pert_chr == gene_chr)  %>%
    filter(ctrl_perturbation == FALSE) %>% 
    filter(significant  == TRUE) %>%
    filter(enhancer_like_interaction==TRUE) %>%
    filter(abs(dist_to_tss) <=1000000)
  
# extract features for significant ETPs
features <- c("log_2_fold_change")

etp_feat <- etp %>% 
  select(response_id, grna_target, interaction_type, significant, dist_to_tss, ctrl_perturbation,
         all_of(features)) %>% 
  pivot_longer(cols = all_of(features), names_to = "feature", values_to = "value") %>% 
  mutate(dist_to_tss = abs(dist_to_tss))

# reformat feature names for plot
etp_feat <- etp_feat %>% 
  mutate(feature = case_when(
    feature == "log_2_fold_change" ~ "Effect size",
    feature == "hic_int_freq" ~ "Hi-C 3D contact",
    TRUE ~ sub(".+_(.+)", "\\1", feature)
  ))

# convert classes to factors for correct ordering
class_levels <- c("promoter-proximal", "promoter-distal", "intragenic-proximal",
                  "intragenic-distal","intergenic-distal")
etp_feat <- etp_feat %>% 
  mutate(interaction_type = factor(interaction_type, levels = rev(class_levels)))

# extract positive hits only for plots
etp_feat_sig <- filter(etp_feat, significant == TRUE & ctrl_perturbation==FALSE)

# add effect size as separate column
etp_feat_sig <- etp_feat_sig %>% 
  filter(feature == "Effect size") %>% 
  select(response_id, grna_target, effect_size = value) %>%
  left_join(etp_feat_sig, ., by = c("response_id", "grna_target"))

# add group for coloring based on element location
etp_feat_sig <- mutate(etp_feat_sig, class_color = interaction_type)
 

```


```{r}
# colors for each class
class_colors <- c(
  "promoter-proximal" = "#0263a3",        #9ad8fc
  "promoter-distal" = "#9ad8fc",  #6ebff5 
  "intragenic-proximal" ="#c62828",#"#daa520",
  "intragenic-distal" = "#e57373", #f2c94c",
  "intergenic-distal"= "goldenrod" #daa520"
)
```

create full network 
```{r}

es_dat <- filter(etp_feat_sig, feature == "Effect size")

# Unique CRE–gene–class pairs from es_dat
es_pairs <- es_dat %>%
  distinct(
    es_grna_target   = grna_target,
    es_response_id   = response_id,
    interaction_type,
    effect_size
  )

# edges1: CRE -> primary gene
edges1 <- es_pairs %>%
  transmute(
    from       = es_grna_target,
    to         = es_response_id,
    log2fc     = effect_size,  # or abs(effect_size) later for width
    edge_type  = "CRE_to_gene",
   edge_class = as.character(interaction_type),
    edge_color = class_colors[as.character(interaction_type)]
  )
# join es_pairs (primary gene) to discovery_P (grna_target)
matches <- es_pairs %>%
  inner_join(
    discovery_P %>%
      select(grna_target, gene_symbol, log_2_fold_change),
    by = c("es_response_id" = "grna_target")
  )

# edges2: primary gene -> downstream gene_symbol
edges2 <- matches %>%
  transmute(
    from       = es_response_id,
    to         = gene_symbol,
    log2fc     = log_2_fold_change,
    edge_type  = "gene_to_gene",
    edge_class = NA_character_,
    edge_color = NA_character_
  )


# map_dp <- discovery_P %>%
#   select(grna_target, gene_symbol, log_2_fold_change)
# 
# frontier <- matches %>%
#   select(from = gene_symbol) %>%   # start from downstream genes
#   distinct()
# 
# all_follow_edges <- tibble(
#   from      = character(),
#   to        = character(),
#   log2fc    = numeric(),
#   edge_type = character()
# )
# 
# seen <- frontier$from
# 
# repeat {
#   step <- frontier %>%
#     inner_join(map_dp, by = c("from" = "grna_target")) %>%
#     transmute(
#       from,
#       to        = gene_symbol,
#       log2fc    = log_2_fold_change,
#       edge_type = "dp_chain"
#     ) %>%
#     distinct()
# 
#   if (nrow(step) == 0) break
# 
#   all_follow_edges <- bind_rows(all_follow_edges, step)
# 
#   new_nodes <- setdiff(step$to, seen)
#   if (length(new_nodes) == 0) break
# 
#   frontier <- tibble(from = new_nodes)
#   seen <- union(seen, new_nodes)
# }

# recursive downstream chains
#edges3_recursive <- all_follow_edges %>%
#  mutate(
#    edge_class = NA_character_,
#    edge_color = NA_character_
#  )

map_dp <- discovery_P %>%
  select(grna_target, gene_symbol, log_2_fold_change)

frontier <- matches %>%
  select(from = gene_symbol) %>%   # start from downstream genes
  distinct()

all_follow_edges <- tibble(
  from      = character(),
  to        = character(),
  log2fc    = numeric(),
  edge_type = character(),
  depth     = integer()            # optional: track hop number
)

seen <- frontier$from

max_steps <- 1

for (d in seq_len(max_steps)) {

  step <- frontier %>%
    inner_join(map_dp, by = c("from" = "grna_target")) %>%
    transmute(
      from,
      to        = gene_symbol,
      log2fc    = log_2_fold_change,
      edge_type = "dp_chain",
      depth     = d
    ) %>%
    distinct()

  if (nrow(step) == 0) break

  all_follow_edges <- bind_rows(all_follow_edges, step)

  new_nodes <- setdiff(step$to, seen)
  if (length(new_nodes) == 0) break

  frontier <- tibble(from = new_nodes)
  seen <- union(seen, new_nodes)
}

# limited downstream chains (<= 3 hops)
edges3_limited <- all_follow_edges %>%
  select(-depth) %>%               # keep depth if you want it later
  mutate(
    edge_class = NA_character_,
    edge_color = NA_character_
  )
####until here for iterative


#edge_list <- bind_rows(edges1, edges2, edges3_recursive) %>%
edge_list <- bind_rows(edges1, edges2, edges3_limited) %>%
  group_by(from, to) %>%
  summarise(
    log2fc     = first(log2fc),
    edge_type  = paste(unique(edge_type), collapse = ";"),
    edge_class = first(na.omit(edge_class)),
    edge_color = first(na.omit(edge_color)),
    .groups    = "drop"
  )

nodes <- tibble(name = unique(c(edge_list$from, edge_list$to))) %>%
  mutate(
    type = case_when(
      name %in% es_pairs$es_grna_target ~ "CRE",
      TRUE                              ~ "gene"
    )
  )

# Identify self-targeting genes in discovery_P
self_target_genes <- discovery_P %>%
  filter(grna_target == gene_symbol) %>%
  pull(gene_symbol) %>%
  unique()

# Update node types
nodes <- tibble(name = unique(c(edge_list$from, edge_list$to))) %>%
  mutate(
    type = case_when(
      name %in% es_pairs$es_grna_target ~ "CRE",          # CREs
      name %in% pos_neg_genes           ~ "pos_neg_gene", # positive controls: sig & negative
      TRUE                              ~ "gene"          # all other genes
    )
  )

edge_list <- edge_list %>%
  filter(from != to)

g <- graph_from_data_frame(edge_list, directed = TRUE, vertices = nodes)

# Edge colors: class_colors for CRE->gene, grey otherwise
E(g)$color <- ifelse(!is.na(E(g)$edge_color), E(g)$edge_color, "grey60")

# Edge widths: scale |log2fc|, but note edges1 now use effect_size
E(g)$width <- rescale(abs(E(g)$log2fc), to = c(1, 6))

color_map <- c(
  CRE          = "#c03a30",
  gene         = "#848484",
  pos_neg_gene = "#334068"
)



V(g)$shape <- "circle"
V(g)$color <- color_map[V(g)$type]
V(g)$size  <- 3

plot(
  g,
  vertex.label  = NA,
  edge.arrow.size = 0.01,
  layout = layout_with_fr(g)
)
```

filter for only those networks that have at least 1 CRE, and for those genes that have an effect on itself (control genes), but keep non-control genes when its an end node. 
```{r}
## 1. Split edges into CRE→gene vs gene→gene -------------------------------

# assumes you set edge_type when building the graph:
#   "CRE_to_gene"  for CRE -> primary gene
#   "gene_to_gene" and "dp_chain" for gene -> gene edges

is_cre_edge <- E(g)$edge_type == "CRE_to_gene"

cre_gene_edges  <- E(g)[is_cre_edge]         # keep ALL of these
non_cre_edges   <- E(g)[!is_cre_edge]        # gene→gene edges (and similar)


## 2. From gene→gene edges, keep only those where the source is green -----

gene_edges_keep <- non_cre_edges[0]  # empty by default

if (length(non_cre_edges) > 0) {
  # ends() gives from/to vertex indices if names = FALSE
  non_cre_ends <- ends(g, non_cre_edges, names = FALSE)
  from_idx     <- non_cre_ends[, 1]

  # assume green genes were tagged as type == "pos_neg_gene"
  is_from_green <- V(g)$type[from_idx] == "pos_neg_gene"

  gene_edges_keep <- non_cre_edges[is_from_green]
}

# all edges we want to keep
edges_keep <- c(cre_gene_edges, gene_edges_keep)


## 3. Build subgraph from these edges, dropping isolated nodes ------------

g_edges <- subgraph.edges(g, e = edges_keep, delete.vertices = TRUE)


## 4. Keep only components with at least one CRE AND at least one gene ----

comp <- components(g_edges)

comp_has_cre <- tapply(V(g_edges)$type == "CRE",
                       comp$membership, any)

comp_has_gene <- tapply(V(g_edges)$type %in% c("gene", "pos_neg_gene"),
                        comp$membership, any)

comps_to_keep <- which(comp_has_cre & comp_has_gene)

verts_keep <- V(g_edges)[comp$membership %in% comps_to_keep]

g_final <- induced_subgraph(g_edges, vids = verts_keep)


## 5. Plot -----------------------------------------------------------------

# Kamada-Kawai – often the most balanced
#coords <- layout_with_kk(g_final)

# Graphopt – good for large networks
#coords <- layout_with_graphopt(g_final)

coords <-layout_with_fr(g_final)


# DrL – scalable for big graphs, hierarchical feel
#coords <- layout_with_drl(g_final)

plot(
  g_final,
  layout            = coords,
  vertex.label      = NA,
  vertex.size       = V(g_final)$size,
  vertex.color      = V(g_final)$color,
  vertex.frame.color = NA,
  edge.color        = E(g_final)$color,
  edge.width        = E(g_final)$width,
  edge.arrow.size   = 0.01
)

## 3. Centrality & core genes (on g_clean!) ----
# centrality
V(g_final)$degree      <- degree(g_final, mode = "all")
V(g_final)$betweenness <- betweenness(g_final, directed = TRUE, normalized = TRUE)
V(g_final)$eigen       <- evcent(g_final, directed = TRUE)$vector

# genes only (exclude CREs)
gene_nodes <- V(g_final)[V(g_final)$type %in% c("gene", "pos_neg_gene")]

# robust top 10% by degree: pick an explicit number of nodes
n_genes   <- length(gene_nodes)
n_core    <- max(1, ceiling(0.3 * n_genes))  # at least 1
ord_deg   <- order(gene_nodes$degree, decreasing = TRUE, na.last = NA)
core_genes <- gene_nodes[ ord_deg[seq_len(n_core)] ]

## 4. Labels only for core genes ----
V(g_final)$label <- NA                     # clear all labels
V(g_final)$label[core_genes] <- V(g_final)$name[core_genes]  # label only cores

plot(
  g_final,
  layout               = coords,

  # vertices
  vertex.label         = V(g_final)$label,
  vertex.label.cex     = 0.7,
  vertex.label.color   = "black",
  vertex.label.family  = "Arial",
  vertex.label.degree  = -pi / 2,   # label below node
  vertex.label.dist    = 0.5,

  vertex.size          = V(g_final)$size,
  vertex.color         = V(g_final)$color,
  vertex.frame.color   = NA,

  # edges
  edge.color           = E(g_final)$color,
  edge.width           = E(g_final)$width,
  edge.arrow.size      = 0.01
)
```


keep only networks that include both CRE->gene and gene->gene connections. 
```{r}
## 1–2. Directional filter: keep only components with CRE->gene->gene ----
comp <- components(g_final, mode = "weak")  # component membership ignoring direction
comp_ids <- sort(unique(comp$membership))

comp_keep <- sapply(comp_ids, function(cid) {
  vids <- V(g_final)[comp$membership == cid]
  sg   <- induced_subgraph(g_final, vids = vids)

  # CRE -> gene edges
  e_cre <- E(sg)[E(sg)$edge_type == "CRE_to_gene"]
  if (length(e_cre) == 0) return(FALSE)

  # genes that are targets of CREs (middle node in CRE->gene->gene)
  middle_genes <- unique(ends(sg, e_cre, names = FALSE)[, 2])

  # gene -> gene edges
  e_gg <- E(sg)[E(sg)$edge_type %in% c("gene_to_gene", "dp_chain")]
  if (length(e_gg) == 0) return(FALSE)

  gg_from <- ends(sg, e_gg, names = FALSE)[, 1]

  # directional requirement: some CRE-targeted gene is the SOURCE of a gene->gene edge
  any(gg_from %in% middle_genes)
})

comps_keep <- comp_ids[comp_keep]
verts_keep <- V(g_final)[comp$membership %in% comps_keep]

g_clean <- induced_subgraph(g_final, vids = verts_keep)


## 3. Centrality & core genes (on g_clean) ----
V(g_clean)$degree      <- degree(g_clean, mode = "all")
V(g_clean)$betweenness <- betweenness(g_clean, directed = TRUE, normalized = TRUE)
V(g_clean)$eigen       <- evcent(g_clean, directed = TRUE)$vector

# genes only (exclude CREs)
gene_nodes <- V(g_clean)[V(g_clean)$type %in% c("gene", "pos_neg_gene")]

# robust core selection (safe against NA degrees / empty sets)
core_genes <- V(g_clean)[0]  # default empty

if (length(gene_nodes) > 0) {
  ord_deg <- order(gene_nodes$degree, decreasing = TRUE, na.last = NA)
  n_rankable <- length(ord_deg)

  if (n_rankable > 0) {
    n_core <- max(1, ceiling(1 * length(gene_nodes)))  # your setting
    n_core <- min(n_core, n_rankable)                  # critical fix
    core_genes <- gene_nodes[ ord_deg[seq_len(n_core)] ]
  }
}


## 4. Labels only for core genes ----
V(g_clean)$label <- NA
# V(g_clean)$label[core_genes] <- V(g_clean)$name[core_genes]


## 5. Plot ----
coords <- layout_with_fr(g_clean, niter = 3000)

plot(
  g_clean,
  layout               = coords,

  vertex.label         = V(g_clean)$label,
  vertex.label.cex     = 0.7,
  vertex.label.color   = "black",
  vertex.label.family  = "Arial",
  vertex.label.degree  = -pi / 2,
  vertex.label.dist    = 0.5,

  vertex.size          = V(g_clean)$size,
  vertex.color         = V(g_clean)$color,
  vertex.frame.color   = NA,

  edge.color           = E(g_clean)$color,
  edge.width           = E(g_clean)$width,
  edge.arrow.size      = 0.01
)

```

remove end nodes (e.g. genes with only 1 incoming connection and no outgoing connection). not itterative
```{r}
## --- ONE-PASS filter: create g_clean2 (do NOT modify g_clean) ----
# one-pass: drop nodes with ONLY 1 incoming connection (and no outgoing)
indeg  <- degree(g_clean, mode = "in")
outdeg <- degree(g_clean, mode = "out")

verts_drop <- which(indeg == 1 & outdeg == 0)

g_clean2 <- delete_vertices(g_clean, verts_drop)

## --- Centrality & core genes (on g_clean2) ----
V(g_clean2)$degree      <- degree(g_clean2, mode = "all")
V(g_clean2)$betweenness <- betweenness(g_clean2, directed = TRUE, normalized = TRUE)
V(g_clean2)$eigen       <- evcent(g_clean2, directed = TRUE)$vector

gene_nodes <- V(g_clean2)[V(g_clean2)$type %in% c("gene", "pos_neg_gene")]

n_genes <- length(gene_nodes)
n_core  <- max(1, ceiling(1 * n_genes))  # keeping your setting

ord_deg     <- order(gene_nodes$degree, decreasing = TRUE, na.last = NA)
core_genes2 <- gene_nodes[ ord_deg[seq_len(n_core)] ]

## --- Labels only for core genes (on g_clean2) ----
V(g_clean2)$label <- NA
# V(g_clean2)$label[core_genes2] <- V(g_clean2)$name[core_genes2]

## --- Plot g_clean2 ----
coords2 <- layout_with_fr(g_clean2, niter = 3000)

plot(
  g_clean2,
  layout               = coords2,

  # vertices
  vertex.label         = V(g_clean2)$label,
  vertex.label.cex     = 0.7,
  vertex.label.color   = "black",
  vertex.label.family  = "Arial",
  vertex.label.degree  = -pi / 2,
  vertex.label.dist    = 0.5,

  vertex.size          = V(g_clean2)$size,
  vertex.color         = V(g_clean2)$color,
  vertex.frame.color   = NA,

  # edges
  edge.color           = E(g_clean2)$color,
  edge.width           = E(g_clean2)$width,
  edge.arrow.size      = 0.01
)

```


remove end nodes itterative

```{r}
## --- ITERATIVE filter (starting from g_clean): drop genes with ONLY 1 incoming and 0 outgoing ----
drop_one_in_only_iter <- function(g,
                                 gene_types = c("gene", "pos_neg_gene"),
                                 remove_non_genes = FALSE,
                                 verbose = TRUE) {

  step <- 0
  repeat {
    step <- step + 1

    indeg  <- degree(g, mode = "in")
    outdeg <- degree(g, mode = "out")

    is_gene <- V(g)$type %in% gene_types

    if (remove_non_genes) {
      # apply rule to all nodes
      to_drop <- which(indeg == 1 & outdeg == 0)
    } else {
      # apply rule only to genes (recommended)
      to_drop <- which(is_gene & indeg == 1 & outdeg == 0)
    }

    if (length(to_drop) == 0) {
      if (verbose) message("Done after ", step - 1, " iteration(s).")
      break
    }

    if (verbose) message("Iter ", step, ": removing ", length(to_drop), " node(s).")
    g <- delete_vertices(g, to_drop)
  }

  g
}

# Run it (starting from g_clean)
g_clean2_iter <- drop_one_in_only_iter(g_clean, verbose = TRUE)

## --- Centrality (on g_clean2_iter) ----
V(g_clean2_iter)$degree      <- degree(g_clean2_iter, mode = "all")
V(g_clean2_iter)$betweenness <- betweenness(g_clean2_iter, directed = TRUE, normalized = TRUE)
V(g_clean2_iter)$eigen       <- evcent(g_clean2_iter, directed = TRUE)$vector

## --- Labels (same behavior as you had; currently none) ----
V(g_clean2_iter)$label <- NA
# (optional) label genes:
# gene_nodes2 <- V(g_clean2_iter)[V(g_clean2_iter)$type %in% c("gene","pos_neg_gene")]
# V(g_clean2_iter)$label[gene_nodes2] <- V(g_clean2_iter)$name[gene_nodes2]

## --- Plot g_clean2_iter ----
coords2 <- layout_with_fr(g_clean2_iter, niter = 3000)

plot(
  g_clean2_iter,
  layout               = coords2,

  vertex.label         = V(g_clean2_iter)$label,
  vertex.label.cex     = 0.7,
  vertex.label.color   = "black",
  vertex.label.family  = "Arial",
  vertex.label.degree  = -pi / 2,
  vertex.label.dist    = 0.5,

  vertex.size          = V(g_clean2_iter)$size,
  vertex.color         = V(g_clean2_iter)$color,
  vertex.frame.color   = NA,

  edge.color           = E(g_clean2_iter)$color,
  edge.width           = E(g_clean2_iter)$width,
  edge.arrow.size      = 0.01
)

```


remove genes with a single incoming or outgoing edge (itterative)
```{r}
## --- ITERATIVE filter (starting from g_clean):
##     drop genes with (indeg==1 & outdeg==0) OR (indeg==0 & outdeg==1) ----
drop_gene_leaves_iter <- function(g,
                                 gene_types = c("gene", "pos_neg_gene"),
                                 verbose = TRUE) {

  iter <- 0
  repeat {
    iter <- iter + 1

    indeg  <- degree(g, mode = "in")
    outdeg <- degree(g, mode = "out")

    is_gene <- V(g)$type %in% gene_types

    # Rule A: exactly one incoming, zero outgoing
    drop_in_only  <- is_gene & indeg == 1 & outdeg == 0

    # Rule B: exactly one outgoing, zero incoming
    drop_out_only <- is_gene & indeg == 0 & outdeg == 1

    to_drop <- which(drop_in_only | drop_out_only)

    if (length(to_drop) == 0) {
      if (verbose) message("Done after ", iter - 1, " iteration(s).")
      break
    }

    if (verbose) message("Iter ", iter, ": removing ", length(to_drop), " node(s).")
    g <- delete_vertices(g, to_drop)
  }

  g
}

# Run it (starting from g_clean)
g_clean2_iter <- drop_gene_leaves_iter(g_clean, verbose = TRUE)

## --- Centrality (on g_clean2_iter) ----
V(g_clean2_iter)$degree      <- degree(g_clean2_iter, mode = "all")
V(g_clean2_iter)$betweenness <- betweenness(g_clean2_iter, directed = TRUE, normalized = TRUE)
V(g_clean2_iter)$eigen       <- evcent(g_clean2_iter, directed = TRUE)$vector

## --- Labels (same behavior as before; currently none) ----
#V(g_clean2_iter)$label <- NA
# Optional: label genes
 gene_nodes2 <- V(g_clean2_iter)[V(g_clean2_iter)$type %in% c("gene","pos_neg_gene")]
 V(g_clean2_iter)$label[gene_nodes2] <- V(g_clean2_iter)$name[gene_nodes2]

## --- Plot g_clean2_iter ----
coords2 <- layout_with_fr(g_clean2_iter, niter = 3000)

plot(
  g_clean2_iter,
  layout               = coords2,

  vertex.label         = V(g_clean2_iter)$label,
  vertex.label.cex     = 0.7,
  vertex.label.color   = "black",
  vertex.label.family  = "Arial",
  vertex.label.degree  = -pi / 2,
  vertex.label.dist    = 0.5,

  vertex.size          = V(g_clean2_iter)$size,
  vertex.color         = V(g_clean2_iter)$color,
  vertex.frame.color   = NA,

  edge.color           = E(g_clean2_iter)$color,
  edge.width           = E(g_clean2_iter)$width,
  edge.arrow.size      = 0.01
)

```


```{r}

## ------------------------------------------------------------------
## CLEAN + THIN + 3-LAYER LAYOUT (CRE left, primary middle, downstream right)
## starting from: g_clean
## ------------------------------------------------------------------

library(igraph)
library(dplyr)

## 0) Sanity: remove loops (optional but recommended)
g_clean <- simplify(g_clean, remove.loops = TRUE, remove.multiple = FALSE)

## 1) Robustly identify CRE->gene edges (handles concatenated edge_type strings)
is_cre_edge <- grepl("CRE_to_gene", E(g_clean)$edge_type)

## If you accidentally filtered out CRE edges earlier, this will warn you
n_cre_edges <- sum(is_cre_edge)
message("CRE_to_gene edges in g_clean: ", n_cre_edges)
if (n_cre_edges == 0) {
  warning("No CRE_to_gene edges found in g_clean. Check earlier filtering (use grepl not ==).")
}

## 2) Primary targets = 'to' nodes of CRE->gene edges
cre_eids <- E(g_clean)[is_cre_edge]
primary_targets <- unique(ends(g_clean, cre_eids, names = TRUE)[, 2])

message("Primary targets found: ", length(primary_targets))

## 3) Define downstream “end nodes” as sinks among genes (out-degree == 0)
gene_vids <- V(g_clean)[V(g_clean)$type %in% c("gene", "pos_neg_gene")]
gene_sink_names <- gene_vids$name[degree(g_clean, v = gene_vids, mode = "out") == 0]

message("Gene sinks (out-degree==0): ", length(gene_sink_names))

## 4) Keep rule:
##    - keep ALL CRE nodes
##    - keep ALL primary targets
##    - keep ALL sinks (end nodes)
##    - keep downstream nodes within k steps from any primary target
k <- 2  # <- tune this: 1 is very tight, 2 usually good, 3 can blow up fast

if (length(primary_targets) > 0) {
  neigh_k <- ego(g_clean, order = k, nodes = primary_targets, mode = "out")
  downstream_k_names <- unique(V(g_clean)[unlist(neigh_k)]$name)
} else {
  downstream_k_names <- character()
}

keep_names <- unique(c(
  V(g_clean)[V(g_clean)$type == "CRE"]$name,
  primary_targets,
  gene_sink_names,
  downstream_k_names
))

g_layered <- induced_subgraph(g_clean, vids = V(g_clean)[name %in% keep_names])

message("Nodes kept for layered plot: ", vcount(g_layered))
message("Edges kept for layered plot: ", ecount(g_layered))

## 5) Assign 3 layers on the reduced graph:
##    1 = CRE
##    2 = primary target (incoming CRE edge)
##    3 = downstream (everything else)
is_cre_edge2 <- grepl("CRE_to_gene", E(g_layered)$edge_type)
cre_eids2 <- E(g_layered)[is_cre_edge2]
primary2 <- unique(ends(g_layered, cre_eids2, names = TRUE)[, 2])

V(g_layered)$layer <- case_when(
  V(g_layered)$type == "CRE" ~ 1L,
  V(g_layered)$name %in% primary2 ~ 2L,
  TRUE ~ 3L
)

message("Layer counts: ", paste(names(table(V(g_layered)$layer)),
                               table(V(g_layered)$layer),
                               collapse = ", "))

## 6) Build a nice 3-layer layout:
##    - use FR for vertical “structure”
##    - force x = layer
##    - spread y evenly within each layer (prevents stacking)
pad_x   <- 6    # spacing between layers
pad_y   <- 1.2  # vertical spacing within each layer
jitter_y <- 0.10

base <- layout_with_fr(g_layered, niter = 2000)

coords <- base
coords[, 1] <- V(g_layered)$layer * pad_x

coords[, 2] <- ave(
  base[, 2],
  V(g_layered)$layer,
  FUN = function(z) {
    n <- length(z)
    # order by FR y, then place on evenly spaced rails
    ord <- order(z, decreasing = FALSE, na.last = NA)
    y <- ((seq_len(n) - (n + 1) / 2) * pad_y)
    # return in original vertex order
    out <- numeric(n)
    out[ord] <- y
    out
  }
)

# optional tiny jitter so nodes don't sit on perfectly straight lines
set.seed(1)
coords[, 2] <- coords[, 2] + runif(vcount(g_layered), -jitter_y, jitter_y)

## 7) Plot
plot(
  g_layered,
  layout               = coords,

  vertex.label         = V(g_layered)$label,  # keep your existing labels if set; else NA
  vertex.label.cex     = 0.7,
  vertex.label.color   = "black",
  vertex.label.family  = "Arial",
  vertex.label.degree  = -pi / 2,
  vertex.label.dist    = 0.5,

  vertex.size          = V(g_layered)$size,
  vertex.color         = V(g_layered)$color,
  vertex.frame.color   = NA,

  edge.color           = E(g_layered)$color,
  edge.width           = E(g_layered)$width,
  edge.arrow.size      = 0.01,
  edge.curved          = 0.15,

  asp = 0
)

## ------------------------------------------------------------------
## Notes for tuning:
## - If layer 3 is still huge, reduce k to 1.
## - If it's too sparse, increase k to 3 (careful: can explode).
## - Increase pad_y if layer 3 nodes overlap too much.
## - Increase pad_x if layers feel too close.
## ------------------------------------------------------------------


```

 Do the same 3 networks for Stefan's data
 
```{r}
pos_neg_genes_S <- discovery_P_S%>%
  filter(gene_symbol == grna_target, 
    !is.na(log_2_fold_change),
    log_2_fold_change < 0) %>%
  pull(gene_symbol) %>%
  unique()

# join es_pairs (primary gene) to discovery_P_S (grna_target)
matches_S <- es_pairs %>%
  inner_join(
    discovery_P_S %>%
      select(grna_target, gene_symbol, log_2_fold_change),
    by = c("es_response_id" = "grna_target")
  )

edges2_S <- matches_S %>%
  transmute(
    from       = es_response_id,
    to         = gene_symbol,
    log2fc     = log_2_fold_change,
    edge_type  = "gene_to_gene",
    edge_class = NA_character_,
    edge_color = NA_character_
  )

map_dp_S <- discovery_P_S %>%
  select(grna_target, gene_symbol, log_2_fold_change)

frontier_S <- matches_S %>%
  select(from = gene_symbol) %>%   # start from downstream genes
  distinct()

all_follow_edges_S <- tibble(
  from      = character(),
  to        = character(),
  log2fc    = numeric(),
  edge_type = character()
)

seen_S <- frontier_S$from

repeat {
  step_S <- frontier_S %>%
    inner_join(map_dp_S, by = c("from" = "grna_target")) %>%
    transmute(
      from,
      to        = gene_symbol,
      log2fc    = log_2_fold_change,
      edge_type = "dp_chain"
    ) %>%
    distinct()

  if (nrow(step_S) == 0) break

  all_follow_edges_S <- bind_rows(all_follow_edges_S, step_S)

  new_nodes_S <- setdiff(step_S$to, seen_S)
  if (length(new_nodes_S) == 0) break

  frontier_S <- tibble(from = new_nodes_S)
  seen_S <- union(seen_S, new_nodes_S)
}

edges3_recursive_S <- all_follow_edges_S %>%
  mutate(
    edge_class = NA_character_,
    edge_color = NA_character_
  )

edge_list_S <- bind_rows(edges1, edges2_S, edges3_recursive_S) %>%
  group_by(from, to) %>%
  summarise(
    log2fc     = first(log2fc),
    edge_type  = paste(unique(edge_type), collapse = ";"),
    edge_class = first(na.omit(edge_class)),
    edge_color = first(na.omit(edge_color)),
    .groups    = "drop"
  ) %>%
  filter(
    !(from == to & from %in% pos_neg_genes_S)
  )

nodes_S <- tibble(name = unique(c(edge_list_S$from, edge_list_S$to))) %>%
  mutate(
    type = case_when(
      name %in% es_pairs$es_grna_target ~ "CRE",
      TRUE                              ~ "gene"
    )
  )

# Identify self-targeting genes in discovery_P
self_target_genes_S <- discovery_P_S %>%
  filter(grna_target == gene_symbol) %>%
  pull(gene_symbol) %>%
  unique()

# Update node types
nodes_S <- tibble(name = unique(c(edge_list_S$from, edge_list_S$to))) %>%
  mutate(
    type = case_when(
      name %in% es_pairs$es_grna_target ~ "CRE",          # CREs
      name %in% pos_neg_genes_S           ~ "pos_neg_gene", # positive controls: sig & negative
      TRUE                              ~ "gene"          # all other genes
    )
  )

g_S <- graph_from_data_frame(edge_list_S, directed = TRUE, vertices = nodes_S)

# Edge colors: class_colors for CRE->gene, grey otherwise
E(g_S)$color <- ifelse(!is.na(E(g_S)$edge_color), E(g_S)$edge_color, "grey60")

# Edge widths: scale |log2fc|, but note edges1 now use effect_size
E(g_S)$width <- rescale(abs(E(g_S)$log2fc), to = c(1, 6))

color_map <- c(
  CRE          = "#c03a30",
  gene         = "#848484",
  pos_neg_gene = "#334068"
)




V(g_S)$shape <- "circle"
V(g_S)$color <- color_map[V(g_S)$type]
V(g_S)$size  <- 2

plot(
  g_S,
  vertex.label  = NA,
  edge.arrow.size = 0.01,
  layout = layout_with_fr(g_S)
)

```


filter for only those networks that have at least 1 CRE, and for those genes that have an effect on itself (control genes), but keep non-control genes when its an end node. 
```{r}
## 1. Split edges into CRE→gene vs gene→gene -------------------------------

# assumes you set edge_type when building the graph:
#   "CRE_to_gene"  for CRE -> primary gene
#   "gene_to_gene" and "dp_chain" for gene -> gene edges

is_cre_edge_S <- E(g_S)$edge_type == "CRE_to_gene"

cre_gene_edges_S  <- E(g_S)[is_cre_edge_S]         # keep ALL of these
non_cre_edges_S   <- E(g_S)[!is_cre_edge_S]        # gene→gene edges (and similar)


# 2. From gene→gene edges, keep only those where the source is green -----

gene_edges_keep_S <- non_cre_edges_S[0]  # empty by default

if (length(non_cre_edges_S) > 0) {
  # ends() gives from/to vertex indices if names = FALSE
  non_cre_ends_S <- ends(g_S, non_cre_edges_S, names = FALSE)
  from_idx_S     <- non_cre_ends_S[, 1]

  # assume green genes were tagged as type == "pos_neg_gene"
  is_from_green_S <- V(g_S)$type[from_idx_S] == "pos_neg_gene"

  gene_edges_keep_S <- non_cre_edges_S[is_from_green_S]
}

# all edges we want to keep
edges_keep_S <- c(cre_gene_edges_S, gene_edges_keep_S)
#edges_keep_S <- cre_gene_edges_S

## 3. Build subgraph from these edges, dropping isolated nodes ------------

g_edges_S <- subgraph.edges(g_S, e = edges_keep_S, delete.vertices = TRUE)


## 4. Keep only components with at least one CRE AND at least one gene ----

comp_S <- components(g_edges_S)

comp_has_cre_S <- tapply(V(g_edges_S)$type == "CRE",
                       comp_S$membership, any)

comp_has_gene_S <- tapply(V(g_edges_S)$type %in% c("gene", "pos_neg_gene"),
                        comp_S$membership, any)

comps_to_keep_S <- which(comp_has_cre_S & comp_has_gene_S)

verts_keep_S <- V(g_edges_S)[comp_S$membership %in% comps_to_keep_S]

g_final_S <- induced_subgraph(g_edges_S, vids = verts_keep_S)

## 5. Plot -----------------------------------------------------------------

# Kamada-Kawai – often the most balanced
coords <- layout_with_kk(g_final_S)

# Graphopt – good for large networks
#coords <- layout_with_graphopt(g_final_S)



# DrL – scalable for big graphs, hierarchical feel
#coords <- layout_with_drl(g_final_S)

plot(
  g_final_S,
  layout            = coords,
  vertex.label      = NA,
  vertex.size       = V(g_final_S)$size,
  vertex.color      = V(g_final_S)$color,
  vertex.frame.color = NA,
  edge.color        = E(g_final_S)$color,
  edge.width        = E(g_final_S)$width,
  edge.arrow.size   = 0.01
)
```

keep only networks that include both CRE-gene and gene-gene connections. 
```{r}
## 1. Identify gene-to-gene edges (both ends are genes) on g_final ----

E(g_final_S)$gene_gene <- is_gene_gene_edge(g_final_S)

## 2. Filter to components that contain at least one gene→gene edge ----
comp_S <- components(g_final_S)

# map each edge to the component of its source vertex
edge_comp_S <- comp_S$membership[ ends(g_final_S, E(g_final_S), names = FALSE)[,1] ]

comp_has_gene_gene_S <- tapply(E(g_final_S)$gene_gene, edge_comp_S, any)

comps_keep_S  <- as.numeric(names(comp_has_gene_gene_S)[comp_has_gene_gene_S])
verts_keep_S  <- V(g_final_S)[ comp_S$membership %in% comps_keep_S ]

# this is now your cleaned graph
g_clean_S <- induced_subgraph(g_final_S, vids = verts_keep_S)

## 3. Centrality & core genes (on g_clean!) ----
# centrality
V(g_clean_S)$degree      <- degree(g_clean_S, mode = "all")
V(g_clean_S)$betweenness <- betweenness(g_clean_S, directed = TRUE, normalized = TRUE)
V(g_clean_S)$eigen       <- evcent(g_clean_S, directed = TRUE)$vector

# genes only (exclude CREs)
gene_nodes_S <- V(g_clean_S)[V(g_clean_S)$type %in% c("gene", "pos_neg_gene")]

# robust top 10% by degree: pick an explicit number of nodes
n_genes_S   <- length(gene_nodes_S)
n_core_S    <- max(1, ceiling(0.10 * n_genes_S))  # at least 1
ord_deg_S   <- order(gene_nodes_S$degree, decreasing = TRUE, na.last = NA)
core_genes_S <- gene_nodes_S[ ord_deg_S[seq_len(n_core_S)] ]

## 4. Labels only for core genes ----
V(g_clean_S)$label <- NA                     # clear all labels
V(g_clean_S)$label[core_genes_S] <- V(g_clean_S)$name[core_genes_S]  # label only cores
## 5. Plot ----
coords <- layout_with_fr(g_clean_S, niter = 3000)

plot(
  g_clean_S,
  layout               = coords,

  # vertices
  vertex.label         = V(g_clean_S)$label,
  vertex.label.cex     = 0.7,
  vertex.label.color   = "black",
  vertex.label.family  = "Arial",
  vertex.label.degree  = -pi / 2,   # label below node
  vertex.label.dist    = 0.5,

  vertex.size          = V(g_clean_S)$size,
  vertex.color         = V(g_clean_S)$color,
  vertex.frame.color   = NA,

  # edges
  edge.color           = E(g_clean_S)$color,
  edge.width           = E(g_clean_S)$width,
  edge.arrow.size      = 0.05
)
```

